---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# causalMed

**Causal Mediation Analysis with Time-Varying Exposures, Mediators, and Confounders**

`causalMed` implements the **parametric g-formula** for total effect estimation and extends
it with the **survival mediational g-formula** to decompose causal effects into direct and
indirect components. The package supports:

- **Interventional direct and indirect effects** (IDE/IIE) — Lin et al. (2017)
- **Natural direct and indirect effects** (NDE/NIE) — Zheng & van der Laan (2017)

Both approaches handle time-varying exposures, mediators, and confounders in longitudinal
data, including survival outcomes. The standard g-formula component (total effect estimation)
is cross-validated against the `gfoRmula` CRAN package.

## Installation

```r
# Install the development version from GitHub:
# install.packages("devtools")
devtools::install_github("adayim/causalMed")
```

## Key Features

- **G-formula for total effects**: Monte Carlo simulation under user-defined static or
  dynamic (threshold) interventions for binary, continuous, and survival outcomes
- **Mediational g-formula**: decomposes total effects into direct and indirect pathways
  with time-varying mediators and exposure-induced mediator–outcome confounders
- **Two mediation estimands**:
  - `mediation_type = "I"`: interventional IDE/IIE (Lin et al. 2017) — marginal mediator
    distribution via random permutation; does not require cross-world independence
  - `mediation_type = "N"`: natural NDE/NIE (Zheng & van der Laan 2017) — conditional
    mediator distribution with exposure swapping; requires stronger assumptions
- **Flexible model specification**: logistic regression (binary), linear regression
  (normal), multinomial logistic (categorical), and custom simulation functions
- **Recode hooks** (`init_recode`, `in_recode`, `out_recode`) for lag creation,
  cumulative variables, and other within-loop transformations
- **Bootstrap confidence intervals**: non-parametric bootstrap resampling individuals
  (preserving longitudinal structure), with percentile and normal-approximation CIs
- **Parallel bootstrap**: plug in `future::plan(multisession)` on Windows or
  `future::plan(multicore)` on Unix to parallelise across bootstrap replicates

## Usage

### Total effect: standard parametric g-formula

```{r gformula-example}
library(causalMed)

data("nonsurvivaldata", package = "causalMed")

# Manage lagged variables
init_rc <- recodes(lag1_A  = 0,   # At t=0, all lags initialised to 0
                   lag1_L1 = 0,
                   lag1_L2 = 0)

in_rc   <- recodes(lag1_A  = A,   # At each subsequent step, copy current values
                   lag1_L1 = L1,
                   lag1_L2 = L2)

# Specify models in temporal order: L1 → L2 → A → Y
m_L1 <- spec_model(L1    ~ lag1_A + lag1_L1 + V + time,
                   var_type = "normal",  mod_type = "covariate")
m_L2 <- spec_model(L2    ~ lag1_A + lag1_L2 + V + time,
                   var_type = "binary",  mod_type = "covariate")
m_A  <- spec_model(A     ~ lag1_A + L1 + L2 + V + time,
                   var_type = "binary",  mod_type = "exposure")
m_Y  <- spec_model(Y_bin ~ A + L1 + L2,
                   var_type = "binary",  mod_type = "outcome")

models_bin <- list(m_L1, m_L2, m_A, m_Y)

# Define intervention strategies
# NULL  = natural course (draw exposure from its fitted model)
# 1 / 0 = always treat / never treat
ints <- list(natural = NULL, always_treat = 1, never_treat = 0)

# ── 3. Run g-formula ────────────────────────────────────────────────────────
fit_bin <- gformula(
  data        = nonsurvivaldata,
  id_var      = "id",
  time_var    = "time",
  base_vars   = "V",
  exposure    = "A",
  models      = models_bin,
  intervention = ints,
  ref_int     = "natural",
  init_recode = init_rc,
  in_recode   = in_rc,
  mc_sample   = 10000,
  R           = 100,  # set R > 1 for bootstrap CIs; kept low here for speed
  quiet       = TRUE,
  seed        = 2025
)

print(fit_bin)
```

### Mediation analysis: interventional IDE/IIE (Lin et al. 2017)

```{r mediation-example}
library(causalMed)

data("nonsurvivaldata", package = "causalMed")

# Model list must include a mediator model (mod_type = "mediator")
# List order must reflect the temporal DAG: A -> L -> M -> Y
init_med <- recodes(lag1_A = 0, lag1_L1 = 0, lag1_L2 = 0, lag1_M = 0)
in_med   <- recodes(lag1_A = A, lag1_L1 = L1, lag1_L2 = L2, lag1_M = M)

models_med <- list(
  spec_model(A   ~ V + lag1_L1 + lag1_L2 + lag1_A + time,
             var_type = "binary",  mod_type = "exposure"),
  spec_model(L1  ~ V + A + lag1_L1 + time,
             var_type = "normal",  mod_type = "covariate"),
  spec_model(L2  ~ V + A + lag1_L2 + time,
             var_type = "binary",  mod_type = "covariate"),
  spec_model(M   ~ V + A + L1 + L2 + lag1_M + time,
             var_type = "normal",  mod_type = "mediator"),   # <-- mediator
  spec_model(Y_bin ~ V + A + M + L1 + L2,
             var_type = "binary",  mod_type = "outcome")
)

fit_med <- mediation(
  data           = nonsurvivaldata,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y_bin",
  models         = models_med,
  init_recode    = init_med,
  in_recode      = in_med,
  mediation_type = "I",     # Interventional IDE/IIE
  mc_sample      = 10000,
  R              = 100,
  quiet          = TRUE,
  seed           = 2025
)

print(fit_med)
```

The `estimate` component of `fit_med` contains:

| Effect | Definition |
|--------|------------|
| Indirect effect | Q(1,1) − Q(1,0): effect through the mediator pathway |
| Direct effect   | Q(1,0) − Q(0,0): effect not through the mediator |
| Total effect    | Q(1,1) − Q(0,0): sum of direct and indirect |
| Mediation proportion | Indirect / Total × 100% |

where Q(a₁, a₂) is the mean outcome when exposure is set to a₁ and the mediator
distribution is drawn from the distribution it would have had under exposure a₂.

### Natural NDE/NIE (Zheng & van der Laan 2017)

Change one argument:

```{r natural-example, eval = FALSE}
fit_natural <- mediation(
  ...,                     # same arguments as above
  mediation_type = "N"    # natural NDE/NIE (Zheng & van der Laan 2017)
)
```

Natural effects condition the mediator model on the individual's own covariate history
but evaluate it at the alternative exposure level (exposure swapping). They require
stronger sequential no-unmeasured-confounding assumptions than interventional effects.

### Enabling parallel bootstrap

```{r parallel-example, eval = FALSE}
library(future)
plan(multisession)   # Windows; use plan(multicore) on Unix/macOS

fit_parallel <- mediation(..., R = 500)

plan(sequential)     # reset after use
```

## Data format

Input data must be in **long format** with one row per subject per time point. The time
variable should be an ordered integer starting at 0. Variables used in models as lags
(e.g., `A_lag1`) must be pre-computed and present in the data, or created via
`init_recode` / `in_recode` hooks. The `nonsurvivaldata` bundled with the package
illustrates the required structure.

```{r data-example, eval = FALSE}
data("nonsurvivaldata", package = "causalMed")
head(nonsurvivaldata)
#>   id time V          L A         M Y M_lag1    L_lag1 A_lag1
#> 1  1    0 1 -0.4557851 0 0.1847927 0      0  0.000000      0
#> 2  1    1 1  0.4651968 0 0.9559358 0      0 -0.455785      0
#> 3  1    2 1  0.2498987 1 1.3503895 0      1  0.465197      0
```

## References

1. Westreich, D., Cole, S. R., Young, J. G., et al. (2012).
   The parametric g-formula to estimate the effect of HAART on incident AIDS or death.
   *Statistics in Medicine*, 31, 2000–2009.
   [doi:10.1002/sim.5316](https://doi.org/10.1002/sim.5316)

2. McGrath, S., Lin, V., Zhang, Z., et al. (2020).
   gfoRmula: An R Package for Estimating the Effects of Sustained Treatment Strategies
   via the Parametric g-Formula.
   *Patterns*, 1, 100008.
   [doi:10.1016/j.patter.2020.100008](https://doi.org/10.1016/j.patter.2020.100008)

3. Lin, S. H., Young, J. G., Logan, R., & VanderWeele, T. J. (2017).
   Mediation analysis for a survival outcome with time-varying exposures, mediators,
   and confounders.
   *Statistics in Medicine*, 36(26), 4153–4166.
   [doi:10.1002/sim.7426](https://doi.org/10.1002/sim.7426)

4. Zheng, W., & van der Laan, M. (2017).
   Longitudinal mediation analysis with time-varying mediators and exposures, with
   application to survival outcomes.
   *Journal of Causal Inference*, 5(2).
   [doi:10.1515/jci-2016-0006](https://doi.org/10.1515/jci-2016-0006)
