---
output: github_document
---
<!-- badges: start -->
  [![R-CMD-check](https://github.com/adayim/causalMed/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/adayim/causalMed/actions/workflows/R-CMD-check.yaml)
  <!-- badges: end -->
  
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# causalMed

**Causal Mediation Analysis with Time-Varying Exposures, Mediators, and Confounders**

`causalMed` implements the **parametric g-formula** for total effect estimation and extends
it with the **survival mediational g-formula** to decompose causal effects into direct and
indirect components. The package supports:

- **Interventional direct and indirect effects** (IDE/IIE) — Lin et al. (2017)
- **Natural direct and indirect effects** (NDE/NIE) — Zheng & van der Laan (2017)

Both approaches handle time-varying exposures, mediators, and confounders in longitudinal
data, including survival outcomes. The standard g-formula component (total effect estimation)
is cross-validated against the `gfoRmula` CRAN package.

## Installation

```r
# Install the development version from GitHub:
# install.packages("devtools")
devtools::install_github("adayim/causalMed")
```

## Key Features

- **G-formula for total effects**: Monte Carlo simulation under user-defined static or
  dynamic (threshold) interventions for binary, continuous, and survival outcomes
- **Mediational g-formula**: decomposes total effects into direct and indirect pathways
  with time-varying mediators and exposure-induced mediator–outcome confounders
- **Two mediation estimands**:
  - `mediation_type = "I"`: interventional IDE/IIE (VanderWeele & Tchetgen
    Tchetgen 2017; Lin et al. 2017; Yamamuro et al. 2021) — cross-world
    mediator drawn as a **joint M(1:T) trajectory** by row-permuting the
    reference (a*) cohort, matching the SAS `mGFORMULA` macro and the
    algorithm of Yamamuro et al. 2021. Does not require cross-world
    independence. **Multiple mediator models** are supported under this
    estimand (Yamamuro 2021 §4); per-mediator IIE(M_k) is reported
    alongside the overall IDE/TE.
  - `mediation_type = "N"`: natural NDE/NIE (Zheng & van der Laan 2017) — conditional
    mediator distribution with exposure swapping; requires stronger assumptions
    (single mediator only). Natural effects are **not identifiable** when a
    confounder of the mediator–outcome relationship is itself affected by the
    exposure; `mediation()` detects this, warns, and repeats the caveat in
    `print()` — prefer `mediation_type = "I"` in that setting.
- **Two estimators for natural effects**: the parametric g-formula plug-in
  (`estimator = "gcomp"`, the default) or a **targeted maximum likelihood
  estimator** (`estimator = "tmle"`, Zheng & van der Laan 2017 §4.3) — multiply
  robust (consistent when only certain subsets of the nuisance models are
  correct), with Wald confidence intervals from the efficient influence curve,
  so no bootstrap is needed. Available for `mediation_type = "N"` only.
- **Flexible model specification**: logistic regression (binary), linear regression
  (normal), multinomial logistic (categorical), and custom simulation functions.
  Numeric draws are clipped to the observed range of the response by default;
  use `spec_model(truncate = FALSE)` to draw from the untruncated fitted
  distribution.
- **Recode hooks** (`init_recode`, `in_recode`, `out_recode`) for lag creation,
  cumulative variables, and other within-loop transformations
- **Bootstrap confidence intervals**: non-parametric bootstrap resampling individuals
  (preserving longitudinal structure), with percentile and normal-approximation CIs
- **Parallel bootstrap**: plug in `future::plan(multisession)` on Windows or
  `future::plan(multicore)` on Unix to parallelise across bootstrap replicates

## Usage

### Total effect: standard parametric g-formula

```{r gformula-example}
library(causalMed)

data("nonsurvivaldata", package = "causalMed")

# Manage lagged variables
init_rc <- recodes(lag1_A  = 0,   # At t=0, all lags initialised to 0
                   lag1_L1 = 0,
                   lag1_L2 = 0)

in_rc   <- recodes(lag1_A  = A,   # At each subsequent step, copy current values
                   lag1_L1 = L1,
                   lag1_L2 = L2)

# Specify models in temporal order: A → L1 → L2 → Y (exposure first,
# current A in the confounder models — matching the documented DGP)
m_A  <- spec_model(A     ~ V + lag1_A + lag1_L1 + lag1_L2 + time,
                   var_type = "binary",  mod_type = "exposure")
m_L1 <- spec_model(L1    ~ V + A + lag1_L1 + time,
                   var_type = "normal",  mod_type = "covariate")
m_L2 <- spec_model(L2    ~ V + A + lag1_L2 + time,
                   var_type = "binary",  mod_type = "covariate")
m_Y  <- spec_model(Y_bin ~ V + A + L1 + L2,
                   var_type = "binary",  mod_type = "outcome")

models_bin <- list(m_A, m_L1, m_L2, m_Y)

# Define intervention strategies
# NULL  = natural course (draw exposure from its fitted model)
# 1 / 0 = always treat / never treat
ints <- list(natural = NULL, always_treat = 1, never_treat = 0)

# Run g-formula
fit_bin <- gformula(
  data        = nonsurvivaldata,
  id_var      = "id",
  time_var    = "time",
  base_vars   = "V",
  exposure    = "A",
  models      = models_bin,
  intervention = ints,
  ref_int     = "natural",
  init_recode = init_rc,
  in_recode   = in_rc,
  mc_sample   = 10000,
  R           = 100,  # set R > 1 for bootstrap CIs; kept low here for speed
  quiet       = TRUE,
  seed        = 2025
)

print(fit_bin)
```

The list order must match your assumed data-generating process — here the
exposure is decided first within each period and the confounders respond to
it, as documented in `?nonsurvivaldata`.

### Mediation analysis: interventional IDE/IIE (Lin et al. 2017)

```{r mediation-example}
library(causalMed)

data("nonsurvivaldata", package = "causalMed")

# Model list must include a mediator model (mod_type = "mediator")
# List order must reflect the temporal DAG: A -> L -> M -> Y
init_med <- recodes(lag1_A = 0, lag1_L1 = 0, lag1_L2 = 0, lag1_M = 0)
in_med   <- recodes(lag1_A = A, lag1_L1 = L1, lag1_L2 = L2, lag1_M = M)

models_med <- list(
  spec_model(A   ~ V + lag1_L1 + lag1_L2 + lag1_A + time,
             var_type = "binary",  mod_type = "exposure"),
  spec_model(L1  ~ V + A + lag1_L1 + time,
             var_type = "normal",  mod_type = "covariate"),
  spec_model(L2  ~ V + A + lag1_L2 + time,
             var_type = "binary",  mod_type = "covariate"),
  spec_model(M   ~ V + A + L1 + L2 + lag1_M + time,
             var_type = "normal",  mod_type = "mediator"),   # <-- mediator
  spec_model(Y_bin ~ V + A + M + L1 + L2,
             var_type = "binary",  mod_type = "outcome")
)

fit_med <- mediation(
  data           = nonsurvivaldata,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y_bin",
  models         = models_med,
  init_recode    = init_med,
  in_recode      = in_med,
  mediation_type = "I",     # Interventional IDE/IIE
  mc_sample      = 10000,
  R              = 100,
  quiet          = TRUE,
  seed           = 2025
)

print(fit_med)
```

The `estimate` component of `fit_med` contains:

| Effect | Definition |
|--------|------------|
| Indirect effect | Q(1,1) − Q(1,0): effect through the mediator pathway |
| Direct effect   | Q(1,0) − Q(0,0): effect not through the mediator |
| Total effect    | natural plug-in g-formula contrast E[Y₁] − E[Y₀] |
| TE − (Direct + Indirect) | mediated-interaction residual (interventional only; exactly 0, and omitted, for natural effects) |
| Mediation Proportion | (Total − Direct) / Total × 100% (additive) |
| Mediation Proportion (multiplicative) | RR-scale proportion mediated (Lin et al. 2017, Table 2) |

With `R > 1`, `estimate` additionally carries bootstrap SEs and percentile/normal
CIs on the RD and RR scales, and the returned object gains `boot_estimates` (the
per-replicate draws), `data_summary`, and an `observed` nonparametric benchmark
that `print()` displays next to the simulated means.

For `mediation_type = "I"`, Q(a₁, a₂) is the mean outcome when exposure is set to
a₁ and each mediator is a **stochastic draw** from its marginal distribution
under exposure a₂. The direct and indirect effects then sum to the
*interventional overall effect* Q(1,1) − Q(0,0), which generally differs from
the total effect; their gap is the residual row. For `mediation_type = "N"` the
references use each subject's own mediator and the decomposition sums exactly to
the total effect (no residual row).

Under `mediation_type = "I"`, the mediators in **every** intervention — the
references Q(0,0) and Q(1,1) as well as the cross-world Q(1,0) — are drawn as
**joint trajectories**: a natural-course cohort is simulated under each a*, and
the intervention assigns each subject the full M(1:T) of a randomly permuted
reference-cohort individual (each mediator permuted independently). This matches the SAS
`mGFORMULA` macro (Lin et al. 2017, eAppendix) and Yamamuro et al. 2021
(Figure 3, step 3). Mediator values are not survival-weighted; the full
reference cohort is used. See `?mediation` for details.

### Natural NDE/NIE (Zheng & van der Laan 2017)

Change one argument:

```{r natural-example, eval = FALSE}
fit_natural <- mediation(
  ...,                     # same arguments as above
  mediation_type = "N"    # natural NDE/NIE (Zheng & van der Laan 2017)
)
```

Natural effects condition the mediator model on the individual's own covariate history
but evaluate it at the alternative exposure level (exposure swapping). They require
stronger sequential no-unmeasured-confounding assumptions than interventional effects.
In particular, natural direct and indirect effects are **not identifiable** when a
mediator–outcome confounder is itself affected by prior exposure (Avin, Shpitser &
Pearl 2005; VanderWeele & Tchetgen Tchetgen 2017). `mediation()` detects this
pattern, warns at run time, and restates the caveat under the decomposition when
you `print()` the result; use `mediation_type = "I"` in that setting.

#### Targeted estimation (TMLE)

For natural effects only, the plug-in simulation can be replaced by the targeted
maximum likelihood estimator of Zheng & van der Laan (2017, §4.3):

```{r tmle-example, eval = FALSE}
fit_tmle <- mediation(
  ...,                     # same arguments as above
  mediation_type = "N",
  estimator      = "tmle"  # multiply robust; EIC Wald CIs, no bootstrap
)
```

It runs backward targeted regressions instead of forward simulation, so it is
multiply robust (consistent when only certain subsets of the nuisance models are
correct) and reports Wald CIs from the efficient influence curve — `R` and
`mc_sample` are ignored. It requires an exposure model, a binary outcome, and
lag-style recodes only (`recodes(lag_A = A)`; exposure lags must be
first-order); derived recodes such as splines or
cumulative counts need `estimator = "gcomp"`. Inspect any positivity warnings it
reports before trusting the affected quantities.

### Enabling parallel bootstrap

```{r parallel-example, eval = FALSE}
library(future)
plan(multisession)   # Windows; use plan(multicore) on Unix/macOS

fit_parallel <- mediation(..., R = 500)

plan(sequential)     # reset after use
```

## Data format

Input data must be in **long format** with one row per subject per time point. The time
variable should be an ordered integer starting at 0. Variables used in models as lags
(e.g., `lag1_A`) must be pre-computed and present in the data, or created via
`init_recode` / `in_recode` hooks. The `nonsurvivaldata` bundled with the package
illustrates the required structure.

```{r data-example}
data("nonsurvivaldata", package = "causalMed")
head(nonsurvivaldata)
```

## References

1. Westreich, D., Cole, S. R., Young, J. G., et al. (2012).
   The parametric g-formula to estimate the effect of HAART on incident AIDS or death.
   *Statistics in Medicine*, 31, 2000–2009.
   [doi:10.1002/sim.5316](https://doi.org/10.1002/sim.5316)

2. McGrath, S., Lin, V., Zhang, Z., et al. (2020).
   gfoRmula: An R Package for Estimating the Effects of Sustained Treatment Strategies
   via the Parametric g-Formula.
   *Patterns*, 1, 100008.
   [doi:10.1016/j.patter.2020.100008](https://doi.org/10.1016/j.patter.2020.100008)

3. Lin, S. H., Young, J. G., Logan, R., & VanderWeele, T. J. (2017).
   Mediation analysis for a survival outcome with time-varying exposures, mediators,
   and confounders.
   *Statistics in Medicine*, 36(26), 4153–4166.
   [doi:10.1002/sim.7426](https://doi.org/10.1002/sim.7426)

4. Zheng, W., & van der Laan, M. (2017).
   Longitudinal mediation analysis with time-varying mediators and exposures, with
   application to survival outcomes.
   *Journal of Causal Inference*, 5(2).
   [doi:10.1515/jci-2016-0006](https://doi.org/10.1515/jci-2016-0006)

5. VanderWeele, T. J., & Tchetgen Tchetgen, E. J. (2017).
   Mediation analysis with time varying exposures and mediators.
   *Journal of the Royal Statistical Society: Series B*, 79(3), 917–938.
   [doi:10.1111/rssb.12194](https://doi.org/10.1111/rssb.12194)

6. Yamamuro, S., Shinozaki, T., Iimuro, S., & Matsuyama, Y. (2021).
   Mediational g-formula for time-varying treatment and repeated-measured
   multiple mediators: Application to atorvastatin's effect on cardiovascular
   disease via cholesterol lowering and anti-inflammatory actions in elderly
   type 2 diabetics. *Statistical Methods in Medical Research*, 30(8),
   1782–1799.
   [doi:10.1177/09622802211025988](https://doi.org/10.1177/09622802211025988)
