---
output: github_document
---
<!-- badges: start -->
[![R-CMD-check](https://github.com/adayim/causalMed/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/adayim/causalMed/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/adayim/causalMed/graph/badge.svg)](https://app.codecov.io/gh/adayim/causalMed)
<!-- badges: end -->

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# causalMed

**Mediation analysis with time-varying exposures, mediators, and confounders**

`causalMed` uses the parametric g-formula to split the effect of a time-varying
exposure into a direct effect and an indirect effect through one or more
mediators, including when the confounders are themselves affected by earlier
exposure. It estimates:

- **interventional** direct and indirect effects (`mediation_type = "I"`, the
  default; Lin et al. 2017, VanderWeele & Tchetgen Tchetgen 2017), with one or
  more mediators (Yamamuro et al. 2021);
- **natural** direct and indirect effects (`mediation_type = "N"`; Zheng & van
  der Laan 2017), by g-computation or targeted minimum loss-based estimation;
- **total effects** of static or dynamic interventions with `gformula()`,
  checked against the CRAN package `gfoRmula` (McGrath et al. 2020).

Outcomes can be continuous or binary at the end of follow-up, or discrete-time
survival. Confidence intervals come from a subject-level bootstrap, which can
run in parallel.

## Installation

```r
# install.packages("remotes")
remotes::install_github("adayim/causalMed")
```

## Example: interventional direct and indirect effects

`nonsurvivaldata` follows 3,000 subjects over five time points: a baseline
covariate `V`, a binary exposure `A`, confounders `L1` and `L2` that respond to
`A`, a mediator `M`, and a binary outcome `Y_bin` at the end of follow-up.

Each time-varying variable gets a model, listed in the order the variables are
generated within a time point. Lagged variables are created inside the
simulation with `recodes()`: `init_recode` sets them at the first time point and
`in_recode` updates them at each later one.

```{r mediation-example}
library(causalMed)

models <- list(
  spec_model(A ~ V + lag1_A + lag1_L1 + lag1_L2 + time,
             var_type = "binary", mod_type = "exposure"),
  spec_model(L1 ~ V + A + lag1_L1 + time,
             var_type = "normal", mod_type = "covariate"),
  spec_model(L2 ~ V + A + lag1_L2 + time,
             var_type = "binary", mod_type = "covariate"),
  spec_model(M ~ V + A + L1 + L2 + lag1_M + time,
             var_type = "normal", mod_type = "mediator"),
  spec_model(Y_bin ~ V + A + M + L1 + L2,
             var_type = "binary", mod_type = "outcome")
)

fit <- mediation(
  data        = nonsurvivaldata,
  id_var      = "id",
  time_var    = "time",
  base_vars   = "V",
  exposure    = "A",
  outcome     = "Y_bin",
  models      = models,
  init_recode = recodes(lag1_A = 0, lag1_L1 = 0, lag1_L2 = 0, lag1_M = 0),
  in_recode   = recodes(lag1_A = A, lag1_L1 = L1, lag1_L2 = L2, lag1_M = M),
  mc_sample   = 10000,
  R           = 100,   # bootstrap replicates; kept low here for speed
  quiet       = TRUE,
  seed        = 2025
)

fit
```

By default the effects compare always exposed with never exposed; other
static regimes are set with `exposure_regime` and `reference_regime`. The rows
of the decomposition are:

| Row | Meaning |
|-----|---------|
| Direct effect | `Phi10 - Phi00`: effect of the exposure with the mediator drawn from its distribution under no exposure |
| Indirect effect | `Phi11 - Phi10`: with exposure, the effect of shifting the mediator's distribution from its unexposed to its exposed form |
| Total effect | `nat1 - nat0`: the ordinary g-formula total effect |
| TE - (Direct + Indirect) | the decomposition residual: direct + indirect sum to the interventional overall effect `Phi11 - Phi00`, not to the total effect |
| Mediation Proportion | indirect effect as a percentage of the interventional overall effect |

## Other analyses

**Natural effects.** Set `mediation_type = "N"`. Zheng & van der Laan (2017,
Lemma 1) identify these effects under sequential randomization and positivity;
reading them as individual-level natural effects additionally requires a
cross-world assumption that is not expected to hold when a mediator-outcome
confounder is affected by the exposure (Avin, Shpitser & Pearl 2005), as `L1`
and `L2` are here. `mediation()` warns when a covariate model includes the
exposure. `estimator = "tmle"` gives the targeted estimator of Zheng & van der
Laan (2017, Section 4.3); see `?mediation` for what it accepts.

**Total effects.** `gformula()` needs no mediator model, and takes a named list
of static (`1`, `0`) or dynamic (`dyn_int()`) interventions:

```{r gformula-example, eval = FALSE}
models_te <- c(models[1:3],   # exposure and confounder models from above
               list(spec_model(Y_bin ~ V + A + L1 + L2,
                               var_type = "binary", mod_type = "outcome")))

fit_te <- gformula(
  data         = nonsurvivaldata, id_var = "id", time_var = "time",
  base_vars    = "V", exposure = "A", models = models_te,
  intervention = list(natural = NULL, always = 1, never = 0,
                      treat_if_L1_pos = dyn_int(as.numeric(lag1_L1 > 0))),
  init_recode  = recodes(lag1_A = 0, lag1_L1 = 0, lag1_L2 = 0),
  in_recode    = recodes(lag1_A = A, lag1_L1 = L1, lag1_L2 = L2),
  R            = 500
)
```

**Parallel bootstrap.** Set a `future` plan before the call:

```{r parallel-example, eval = FALSE}
future::plan(future::multisession)
fit <- mediation(..., R = 500)
future::plan(future::sequential)
```

## Data format

Data must be in long format, one row per subject per time point, with a numeric
time variable. For a survival outcome, remove every row after the event and
after loss to follow-up. The simulation starts from the subject identifier and
the baseline covariates only, so lags and other derived variables must be
created with the recode hooks even if they already exist in the data.

## Learn more

- `vignette("causalMed-01-overview")`: data format, model specification and
  quick starts.
- `vignette("causalMed-02-mediation")`: the two estimands, survival outcomes,
  multiple mediators, censoring, and natural effects.
- `vignette("causalMed-03-gformula")`: total effects, dynamic interventions,
  custom covariate distributions, and a published replication.
- `vignette("causalMed-04-vs-gfoRmula")`: comparison with `gfoRmula`.

## References

- Avin, C., Shpitser, I., & Pearl, J. (2005). Identifiability of path-specific
  effects. *Proceedings of the 19th International Joint Conference on
  Artificial Intelligence*, 357–363.
- Lin, S. H., Young, J. G., Logan, R., & VanderWeele, T. J. (2017). Mediation
  analysis for a survival outcome with time-varying exposures, mediators, and
  confounders. *Statistics in Medicine*, 36(26), 4153–4166.
  [doi:10.1002/sim.7426](https://doi.org/10.1002/sim.7426)
- McGrath, S., Lin, V., Zhang, Z., et al. (2020). gfoRmula: An R package for
  estimating the effects of sustained treatment strategies via the parametric
  g-formula. *Patterns*, 1, 100008.
  [doi:10.1016/j.patter.2020.100008](https://doi.org/10.1016/j.patter.2020.100008)
- VanderWeele, T. J., & Tchetgen Tchetgen, E. J. (2017). Mediation analysis
  with time varying exposures and mediators. *Journal of the Royal Statistical
  Society: Series B*, 79(3), 917–938.
  [doi:10.1111/rssb.12194](https://doi.org/10.1111/rssb.12194)
- Yamamuro, S., Shinozaki, T., Iimuro, S., & Matsuyama, Y. (2021). Mediational
  g-formula for time-varying treatment and repeated-measured multiple
  mediators. *Statistical Methods in Medical Research*, 30(8), 1782–1799.
  [doi:10.1177/09622802211025988](https://doi.org/10.1177/09622802211025988)
- Zheng, W., & van der Laan, M. (2017). Longitudinal mediation analysis with
  time-varying mediators and exposures, with application to survival outcomes.
  *Journal of Causal Inference*, 5(2).
  [doi:10.1515/jci-2016-0006](https://doi.org/10.1515/jci-2016-0006)
