---
title: "Getting Started with causalMed"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    highlight: null
vignette: >
  %\VignetteIndexEntry{Getting Started with causalMed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Introduction

**causalMed** implements the parametric g-formula for two related goals:

1. **Total effect estimation** (`gformula`): estimate the counterfactual mean outcome had everyone in the study followed a given exposure strategy, using the standard parametric g-formula of Westreich et al. (2012) and McGrath et al. (2020).

2. **Causal mediation analysis** (`mediation`): decompose the total effect into a direct component (not through the mediator) and an indirect component (through the mediator), using either:
   - **Interventional direct/indirect effects (IDE/IIE)** via the survival mediational g-formula of Lin et al. (2017);
   - **Natural direct/indirect effects (NDE/NIE)** via the longitudinal identification formula of Zheng & van der Laan (2017).

Both approaches handle time-varying exposures, mediators, and confounders — including confounders that are themselves affected by prior exposure (exposure-induced mediator–outcome confounding), a setting where standard regression-based mediation fails.

```{r load}
library(causalMed)
library(data.table)
```

---

## Data Structure

The package requires **long-format** data: one row per subject per time point.

```{r data}
data("nonsurvivaldata")
head(nonsurvivaldata, 10)
```

The `nonsurvivaldata` dataset contains 1 000 subjects observed at five time points (0, 1, 2, 3, 4):

| Variable | Role |
|----------|------|
| `id` | Subject identifier |
| `time` | Time index (0, 1, 2, 3, 4) |
| `V` | Time-fixed baseline covariate |
| `A` | Time-varying binary exposure |
| `L1` | Time-varying continuous confounder |
| `L2` | Time-varying binary confounder |
| `M` | Time-varying continuous mediator |
| `Y_bin` | Binary outcome (end-of-follow-up) |

The assumed temporal ordering within each period is **A → L → M → Y**, meaning that within any time point, exposure precedes the confounders, which precede the mediator, which precedes the outcome (this matches the data-generating process documented in `?nonsurvivaldata`).

---

## Specifying Models

Every variable that is not time-fixed and not the identifier needs a parametric model. Models are created with `spec_model()` and collected into a list **in the temporal order they should be simulated**.

```{r spec-model-args, eval=FALSE}
spec_model(
  formula,            # Standard R formula: response ~ predictors
  var_type,           # Distribution for simulation
  mod_type,           # Role in the causal structure
  subset  = NULL,     # Optional condition restricting which rows are used
  recode  = NULL      # Optional within-loop recoding (see recodes())
)
```

### `var_type` — how to draw simulated values

| Value | Distribution | Typical use |
|-------|-------------|-------------|
| `"binary"` | logistic regression | Binary covariates, exposure, outcome |
| `"normal"` | linear regression | Continuous covariates, mediator |
| `"categorical"` | Multinomial via `nnet::multinom` | Unordered categorical covariates |
| `"custom"` | User-supplied function | Specialised distributions |

### `mod_type` — causal role

| Value | Description |
|-------|-------------|
| `"covariate"` | Time-varying confounder |
| `"exposure"` | Intervention variable |
| `"mediator"` | Mediator (required for `mediation()`) |
| `"outcome"` | Binary end-of-follow-up outcome |
| `"survival"` | Discrete-time event indicator (hazard model) |
| `"censor"` | Right-censoring indicator |

### Temporal ordering is critical

The list order determines the simulation sequence at each time step. This should follow the data-generating mechanism. Confounders at time *t* that are affected by exposure at time *t* must come **after** the exposure model but **before** the outcome model.

---

## Managing Lagged Variables

Most models condition on previous-time values (lags). These are created using `recodes()`, which captures expressions to be evaluated inside the simulated dataset at each step.

Three hooks are available:

| Hook | When applied | Typical use |
|------|-------------|-------------|
| `init_recode` | Once at the first time point | Set lag variables to their baseline values |
| `in_recode` | Start of each subsequent time step | Update lags from the previous step's values |
| `out_recode` | End of each time step | Post-simulation transforms (cumulative sums, etc.) |

```{r recodes-example}
init_rc <- recodes(lag1_A  = 0,   # At t=0, all lags initialised to 0
                   lag1_L1 = 0,
                   lag1_L2 = 0)

in_rc   <- recodes(lag1_A  = A,   # At each subsequent step, copy current values
                   lag1_L1 = L1,
                   lag1_L2 = L2)
```

If lag columns are already present in the data (pre-computed before calling `gformula()`), you do not need recoding hooks.

---

## Total Effect Estimation with `gformula()`

### Example 1 — Binary end-of-follow-up outcome

```{r gformula-binary}
# ── 1. Specify models in temporal order: L1 → L2 → A → Y ──────────────────
m_L1 <- spec_model(L1    ~ lag1_A + lag1_L1 + V + time,
                   var_type = "normal",  mod_type = "covariate")
m_L2 <- spec_model(L2    ~ lag1_A + lag1_L2 + V + time,
                   var_type = "binary",  mod_type = "covariate")
m_A  <- spec_model(A     ~ lag1_A + L1 + L2 + V + time,
                   var_type = "binary",  mod_type = "exposure")
m_Y  <- spec_model(Y_bin ~ A + L1 + L2,
                   var_type = "binary",  mod_type = "outcome")

models_bin <- list(m_L1, m_L2, m_A, m_Y)

# ── 2. Define intervention strategies ──────────────────────────────────────
# NULL  = natural course (draw exposure from its fitted model)
# 1 / 0 = always treat / never treat
ints <- list(natural = NULL, always_treat = 1, never_treat = 0)

# ── 3. Run g-formula ────────────────────────────────────────────────────────
fit_bin <- gformula(
  data        = nonsurvivaldata,
  id_var      = "id",
  time_var    = "time",
  base_vars   = "V",
  exposure    = "A",
  models      = models_bin,
  intervention = ints,
  ref_int     = "natural",
  init_recode = init_rc,
  in_recode   = in_rc,
  mc_sample   = 10000,
  R           = 1,        # set R > 1 for bootstrap CIs; kept low here for speed
  quiet       = TRUE,
  seed        = 2025
)
```

```{r gformula-binary-results}
# Risk (mean outcome) under each strategy
fit_bin$effect_size

# Contrasts vs the reference (natural course)
fit_bin$estimate
```

The `effect_size` table gives the estimated mean outcome under each intervention. The `estimate` table gives contrasts (risk difference and risk ratio) against the reference intervention.

### Example 2 — Survival (time-to-event) outcome

For survival outcomes use `mod_type = "survival"`. The model estimates the discrete-time hazard at each time point; the package accumulates these into cumulative incidence (1 − product-limit survival).

```{r gformula-survival}
data("survivaldata")

m_L  <- spec_model(L ~ V + lag1_L + time,
                   var_type = "normal", mod_type = "covariate")
m_A2 <- spec_model(A ~ V + lag1_A + lag1_L + L + time,
                   var_type = "binary", mod_type = "exposure")
m_Y2 <- spec_model(Y ~ lag1_A + A + L + lag1_L + time,
                   var_type = "binary", mod_type = "survival")  # <-- survival

models_surv <- list(m_L, m_A2, m_Y2)

fit_surv <- gformula(
  data        = survivaldata,
  id_var      = "id",
  base_vars   = "V",
  exposure    = "A",
  time_var    = "time",
  models      = models_surv,
  intervention = list(natural = NULL, never = 0, always = 1),
  ref_int     = "natural",
  init_recode = recodes(lag1_L = 0, lag1_A = 0),
  in_recode   = recodes(lag1_L = L, lag1_A = A),
  mc_sample   = 10000,
  R           = 1,
  quiet       = TRUE,
  seed        = 2025
)

fit_surv$effect_size   # Cumulative incidence by strategy
fit_surv$estimate      # Risk contrasts
```

### A published example — preventing GvHD (Keil et al. 2014)

The toy examples above isolate one feature at a time. A realistic analysis
usually combines several. The package ships `gvhd`, the person-day bone-marrow
transplant data used in the parametric g-formula illustration of Keil et al.
(2014), and the code below reproduces that analysis: the counterfactual risk of
death by day 1825 had graft-versus-host disease (GvHD) **never** occurred,
versus the natural course.

This single example exercises features the toy examples do not:

- **five models** in temporal order — relapse → platelet recovery → GvHD
  (exposure) → censoring → death (survival hazard);
- **absorbing states** via `subset =` (each state is modelled only among those
  not yet in it) plus an `out_recode` carry-forward that locks the state at 1
  afterwards;
- a **censoring model** (`mod_type = "censor"`);
- **restricted cubic splines** of age and day, and day counters, built through
  the three recode hooks working together;
- real daily-scale survival data (137 subjects, 1 825 days).

The models follow Appendix 2 of the paper (see `?gvhd`). Because the response
distributions are all binary/logistic, every model uses `var_type = "binary"`:

```{r gvhd-models, eval=FALSE}
data("gvhd")

# Baseline transforms: restricted cubic splines of age (agecurs1/2) and day
# (daycurs1/2), plus agesq/daysq. The full spline expressions use the knots
# given in the Appendix of Keil et al. (2014).
gvhd <- within(gvhd, {
  agesq <- age^2; daysq <- day^2
  agecurs1 <- (age > 17) * (age - 17)^3 -
              ((age > 30) * (age - 30)^3) * (41.4 - 17) / (41.4 - 30)
  agecurs2 <- (age > 25.4) * (age - 25.4)^3 -
              ((age > 41.4) * (age - 41.4)^3) * (41.4 - 25.4) / (41.4 - 30)
  daycurs1 <- (day > 83.6) * ((day - 83.6) / 83.6)^3 -            # + higher-knot
              (day > 947) * ((day - 947) / 83.6)^3 * (1862.2 - 83.6) / (1862.2 - 947)
  daycurs2 <- (day > 401.4) * ((day - 401.4) / 83.6)^3 -          # terms; see script
              (day > 947) * ((day - 947) / 83.6)^3 * (1862.2 - 401.4) / (1862.2 - 947)
})

models_gvhd <- list(
  # Each time-varying state is absorbing: modelled only while it is still 0.
  spec_model(relapse ~ all + cmv + male + age + gvhdm1 + daysgvhd + platnormm1 +
               daysnoplatnorm + agecurs1 + agecurs2 + day + daysq + wait,
             var_type = "binary", mod_type = "covariate", subset = relapsem1 == 0),
  spec_model(platnorm ~ all + cmv + male + age + agecurs1 + agecurs2 + gvhdm1 +
               daysgvhd + daysnorelapse + wait,
             var_type = "binary", mod_type = "covariate", subset = platnormm1 == 0),
  spec_model(gvhd ~ all + cmv + male + age + platnormm1 + daysnoplatnorm +
               relapsem1 + daysnorelapse + agecurs1 + agecurs2 + day + daysq + wait,
             var_type = "binary", mod_type = "exposure", subset = gvhdm1 == 0),
  spec_model(censlost ~ all + cmv + male + age + daysgvhd + daysnoplatnorm +
               daysnorelapse + agesq + day + daycurs1 + daycurs2 + wait,
             var_type = "binary", mod_type = "censor"),
  spec_model(d ~ all + cmv + male + age + gvhd + platnorm + daysnoplatnorm +
               relapse + daysnorelapse + agesq + wait +
               day * gvhd + daycurs1 * gvhd + daycurs2 * gvhd,
             var_type = "binary", mod_type = "survival")
)
```

The three recode hooks cooperate: `init_recode` seeds day 1 (states at 0,
counters at 0, splines computed), `in_recode` refreshes the functions of day and
the one-day lags at the start of each day, and `out_recode` advances the day
counters and enforces the absorbing carry-forward at the end of each day.

```{r gvhd-recodes, eval=FALSE}
# init_recode / in_recode also recompute daysq, daycurs1, daycurs2 each step
# (omitted here for brevity — identical spline expressions as above).
init_recode <- recodes(daysq = day^2, # + daycurs1, daycurs2 as above
  relapse = 0, gvhd = 0, platnorm = 0, gvhdm1 = 0, relapsem1 = 0, platnormm1 = 0,
  daysnorelapse = 0, daysnoplatnorm = 0, daysnogvhd = 0,
  daysrelapse = 0, daysplatnorm = 0, daysgvhd = 0)

in_recode <- recodes(daysq = day^2, # + daycurs1, daycurs2 as above
  platnormm1 = platnorm, relapsem1 = relapse, gvhdm1 = gvhd)

out_recode <- recodes(
  daysnorelapse  = ifelse(relapse == 0,  daysnorelapse + 1,  daysnorelapse),
  daysnoplatnorm = ifelse(platnorm == 0, daysnoplatnorm + 1, daysnoplatnorm),
  daysnogvhd     = ifelse(gvhd == 0,     daysnogvhd + 1,     daysnogvhd),
  daysgvhd       = ifelse(gvhd == 1,     daysgvhd + 1,       daysgvhd),
  # absorbing carry-forward: once a state was 1 yesterday, keep it at 1
  platnorm = ifelse(platnormm1 == 1, 1, platnorm),
  relapse  = ifelse(relapsem1 == 1,  1, relapse),
  gvhd     = ifelse(gvhdm1 == 1,     1, gvhd))

fit_gvhd <- gformula(gvhd,
  id_var    = "id", time_var = "day", exposure = "gvhd",
  base_vars = c("age", "agesq", "agecurs1", "agecurs2", "male", "cmv", "all", "wait"),
  models    = models_gvhd,
  intervention = list(never = 0),     # a natural-course reference is added automatically
  init_recode = init_recode, in_recode = in_recode, out_recode = out_recode,
  mc_sample = 20000, R = 1, quiet = TRUE, seed = 20260703)

fit_gvhd$effect_size
fit_gvhd$estimate
```

A representative run (`mc_sample = 20000`) gives:

```
#> --- effect_size ---
#>   Intervention   Est
#>        natural 0.612
#>          never 0.588
#>
#> --- estimate (never vs natural) ---
#>   Risk difference  -0.024
#>   Risk ratio        0.961
```

so preventing GvHD is estimated to lower the 5-year risk of death by about
2.4 percentage points. This example is **not evaluated when the vignette is
built** — with 1 825 daily time steps the Monte Carlo loop takes minutes. 
Set `R > 1` there for bootstrap confidence intervals.

### Example 3 — Dynamic (threshold) intervention

A **dynamic intervention** assigns exposure based on a rule that depends on the individual's current covariate values. For example, we might treat individuals when a covariate exceeds a specified threshold. Wrap the rule in `dyn_int()`:

```{r gformula-dynamic}
fit_dyn <- gformula(
  data        = nonsurvivaldata,
  id_var      = "id",
  time_var    = "time",
  base_vars   = "V",
  exposure    = "A",
  models      = models_bin,
  intervention = list(
    natural  = NULL,
    treat_if_L1_pos = dyn_int(as.numeric(L1 > 0))
  ),
  ref_int     = "natural",
  init_recode = init_rc,
  in_recode   = in_rc,
  mc_sample   = 10000,
  R           = 1,
  quiet       = TRUE,
  seed        = 2025
)

fit_dyn$effect_size
fit_dyn$estimate
```

The expression inside `dyn_int()` is evaluated within the simulated dataset at each time step. All current column values — including the exposure after its natural-course draw — are in scope, so rules like `dyn_int(as.numeric(A > 0 & L1 > median(L1)))` are valid.

---

## Causal Mediation Analysis with `mediation()`

The `mediation()` function decomposes the total effect into direct and indirect
components. For **natural** effects (`mediation_type = "N"`) the decomposition is
exact:

$$\underbrace{E[Y_{1,M(1)}] - E[Y_{0,M(0)}]}_{\text{Total effect}} =
  \underbrace{E[Y_{1,M(0)}] - E[Y_{0,M(0)}]}_{\text{Direct effect}} +
  \underbrace{E[Y_{1,M(1)}] - E[Y_{1,M(0)}]}_{\text{Indirect effect}}$$

where $Y_{a, M(a^*)}$ is the potential outcome under exposure $a$ with the
mediator at its natural value under $a^*$.

For **interventional** effects (`mediation_type = "I"`) the mediator is instead
set to a *stochastic* draw $G_{a^*}$ from its marginal distribution under $a^*$
(drawn independently across mediators). The direct and indirect effects are
defined relative to these draws and sum to the **interventional overall effect**
$E[Y_{1,G_1}] - E[Y_{0,G_0}]$, which generally differs from the natural total
effect $E[Y_1] - E[Y_0]$. `mediation()` therefore reports the total effect from a
separate natural-course pass and adds a row `TE - (Direct + Indirect)` — the
mediated-interaction residual ($TE$ minus the overall effect). For natural
effects that residual is exactly zero and the row is omitted.

You must include at least one model with `mod_type = "mediator"` in the model
list, and pass the outcome variable name via `outcome`.

### Requirements and defaults

**Binary exposure.** The exposure variable must be coded as `0` (reference/untreated) and `1` (active/treated). Both Lin et al. (2017) and Zheng & van der Laan (2017) are defined for binary exposures; `mediation()` will stop with an informative error if other values are found.

**One or more mediators.** At least one `mod_type = "mediator"` model is required. Multiple mediators are supported for `mediation_type = "I"` (Yamamuro et al. 2021) — list them in temporal order; `mediation_type = "N"` is single-mediator only.

**Temporal ordering.** The list order determines the simulation sequence. Two orderings are common:

- `A → L → M → S` — confounders at time *t* are not affected by the mediator at time *t* (mediator conditions on L; outcome conditions on A, M, L).
- `A → M → L → S` — confounders at time *t* are affected by both exposure and mediator (Lin et al. 2017 DAG). This is the setting where standard regression fails.

The function checks that exposure precedes the mediator and that the mediator precedes the outcome, and warns if either is violated.

**Joint mediator trajectory (interventional effects).** For `mediation_type = "I"`, a natural-course Monte Carlo cohort is simulated under each treatment level $a^*$ and every individual's full mediator trajectory $M(1{:}T)$ is stored as a row of a pool matrix. Each intervention that fixes a mediator to its $a^*$ value — **including the reference interventions** $\Phi_{00} = E[Y_{0,G_0}]$ and $\Phi_{11} = E[Y_{1,G_1}]$, not only the cross-world $\Phi_{10}$ — row-permutes the relevant pool once and assigns subject $i$ the *entire* trajectory of one randomly chosen pool individual (each mediator permuted independently). This is the joint-trajectory algorithm described by Yamamuro et al. (2021, Figure 3 step 3) and implemented by the SAS `mGFORMULA` macro (Lin et al. 2017 eAppendix); it samples the marginal mediator distribution targeted by Lin et al. (2017, Eq. 4) and VanderWeele & Tchetgen Tchetgen (2017). The pool is not survival-weighted; the full reference cohort is used at every time step (matching both reference SAS implementations).

**Warning summary.** Warnings from model fitting (e.g., convergence, near-separation) are held and printed as a deduplicated summary at function exit. Repeated warnings (e.g., across 500 bootstrap replicates) are shown once with a count.

### Setting up models for mediation

The mediator model must condition on exposure and all confounders at the same time point; the outcome model must include both exposure and mediator. The example below uses the `A → L → M → Y` ordering (confounders not affected by the mediator).

```{r mediation-models}
init_med <- recodes(lag1_A = 0, lag1_L1 = 0, lag1_L2 = 0, lag1_M = 0)
in_med   <- recodes(lag1_A = A, lag1_L1 = L1, lag1_L2 = L2, lag1_M = M)

models_med <- list(
  spec_model(A   ~ V + lag1_L1 + lag1_L2 + lag1_A + time,
             var_type = "binary",  mod_type = "exposure"),
  spec_model(L1  ~ V + A + lag1_L1 + time,
             var_type = "normal",  mod_type = "covariate"),
  spec_model(L2  ~ V + A + lag1_L2 + time,
             var_type = "binary",  mod_type = "covariate"),
  spec_model(M   ~ V + A + L1 + L2 + lag1_M + time,
             var_type = "normal",  mod_type = "mediator"),   # <-- mediator
  spec_model(Y_bin ~ V + A + M + L1 + L2,
             var_type = "binary",  mod_type = "outcome")
)
```

### Interventional effects (IDE/IIE) — Lin et al. (2017); VanderWeele & Tchetgen Tchetgen (2017); Yamamuro et al. (2021)

Interventional effects do not require cross-world independence and are generally preferred for time-varying settings. **Every** decomposition intervention — the references $\Phi_{00}$ and $\Phi_{11}$ as well as the cross-world $\Phi_{10}$ — draws a **joint $M(1{:}T)$ trajectory** from the relevant reference pool (row-permuted once, each mediator independently), breaking the individual-level mediator–confounder dependence while preserving the marginal mediator distribution under $a^*$. The reported total effect is the natural plug-in g-formula contrast (interventions `nat0`/`nat1`), so the direct + indirect effects sum to the interventional *overall* effect rather than the total effect; the gap is the `TE - (Direct + Indirect)` residual.

```{r mediation-interventional}
fit_ide <- mediation(
  data           = nonsurvivaldata,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y_bin",
  models         = models_med,
  init_recode    = init_med,
  in_recode      = in_med,
  mediation_type = "I",     # Interventional IDE/IIE
  mc_sample      = 10000,
  R              = 1,
  quiet          = TRUE,
  seed           = 2025
)
```

```{r mediation-interventional-results}
# Per-intervention risks: interventional Phi00/Phi10/Phi11 plus natural-course nat0/nat1
fit_ide$effect_size

# Effect decomposition (includes the TE - (Direct + Indirect) residual row)
fit_ide$estimate
```

### Natural effects (NDE/NIE) — Zheng & van der Laan (2017)

Natural effects use the **conditional** mediator distribution: at each time step the mediator is predicted using the individual's own covariate history under $a = 1$ but with the exposure argument in the mediator model swapped to $a^* = 0$. This requires the additional assumption of no unmeasured exposure-induced mediator–outcome confounding.

```{r mediation-natural}
fit_nde <- mediation(
  data           = nonsurvivaldata,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y_bin",
  models         = models_med,
  init_recode    = init_med,
  in_recode      = in_med,
  mediation_type = "N",     # Natural NDE/NIE
  mc_sample      = 10000,
  R              = 1,
  quiet          = TRUE,
  seed           = 2025
)

fit_nde$estimate
```

The key identifiability difference between `"I"` and `"N"`:

| | Interventional (`"I"`) | Natural (`"N"`) |
|--|----------------------|-----------------|
| Mediator drawn from | Marginal distribution (permutation) | Conditional distribution (individual history) |
| Cross-world independence required | No | Yes |
| Interpretation | Shift in mediator *population* distribution | Hypothetical individual-level swap |
| Reference | Lin et al. (2017) | Zheng & van der Laan (2017) |

### A targeted (TMLE) alternative for natural effects

The default estimator (`estimator = "gcomp"`) is the parametric g-formula
plug-in: it simulates the counterfactual functionals forward from the fitted
models, so its validity rests on **all** component models being correctly
specified, and confidence intervals require the bootstrap. For
`mediation_type = "N"`, the same paper the plug-in implements (Zheng & van der
Laan 2017, Section 4.3) also derives a **targeted maximum likelihood
estimator**, available via `estimator = "tmle"`:

- **Multiply robust**: the estimate is consistent when specific *subsets* of
  the nuisance models are correct (their Theorem 1 gives three sufficient
  combinations), not only when everything is.
- **Analytic CIs**: 95% Wald intervals come from the efficient influence
  curve, so no bootstrap is needed (`R` is ignored) and there is no Monte
  Carlo simulation layer (`mc_sample` is ignored). A full fit typically takes
  seconds.
- **Requirements**: an exposure model (`mod_type = "exposure"`) is mandatory,
  the outcome must be binary `{0, 1}` (endpoint or survival event indicator),
  and `var_type = "custom"` / `spec_model(subset = )` models are not
  supported. Like all of `mediation_type = "N"`, it is single-mediator only.

```{r mediation-tmle}
fit_tmle <- mediation(
  data           = nonsurvivaldata,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y_bin",
  models         = models_med,
  init_recode    = init_med,
  in_recode      = in_med,
  mediation_type = "N",
  estimator      = "tmle",   # <-- targeted estimator
  quiet          = TRUE,
  seed           = 2025
)

fit_tmle$estimate
```

**Read the warnings.** On this dataset the TMLE reports *practical positivity
violations*: almost no subject in `nonsurvivaldata` remains untreated at
every time point, so the data carry very little direct information about the
"never treated" functional $\Phi_{00}$. The plug-in produces a seemingly
precise answer there by extrapolating from its parametric models; the TMLE
skips the affected targeting steps, warns, and returns an appropriately wide
confidence interval. Neither estimator can conjure information the data do
not contain — but the TMLE tells you when that is happening. When the two
estimators disagree materially, the first things to inspect are these
positivity warnings and the model specifications.

The per-subject influence-curve values are returned in
`fit_tmle$tmle_diag$eic` (one column per functional) for custom contrasts or
diagnostics; `fit_tmle$tmle_diag$eic_mean` should be near zero for
well-supported functionals.

---

## Bootstrap Confidence Intervals

Set `R > 1` to obtain percentile and normal-approximation confidence intervals. The bootstrap resamples whole subjects (all time points together), preserving the longitudinal correlation structure.

```{r bootstrap, eval=FALSE}
fit_boot <- gformula(
  data        = nonsurvivaldata,
  id_var      = "id",
  time_var    = "time",
  base_vars   = "V",
  exposure    = "A",
  models      = models_bin,
  intervention = list(natural = NULL, always = 1),
  ref_int     = "natural",
  init_recode = init_rc,
  in_recode   = in_rc,
  mc_sample   = 10000,
  R           = 200,       # 200 bootstrap replicates
  quiet       = TRUE,
  seed        = 2025
)

# effect_size now includes Sd, perct_lcl/ucl, norm_lcl/ucl
fit_boot$effect_size
fit_boot$estimate

# the individual per-replicate draws are retained in boot_estimates
# ($interventions and $contrasts), for custom intervals or diagnostics
fit_boot$boot_estimates$interventions
```

### Parallel bootstrap

The bootstrap loop uses `future.apply::future_lapply` internally. Enable parallelism by setting a parallel plan before calling the function:

```{r parallel, eval=FALSE}
library(future)
plan(multisession)   # use multiple R sessions in parallel

fit_par <- gformula(..., R = 500)

plan(sequential)     # restore default after use
```

On Unix/macOS, `plan(multicore)` (forking) is more memory-efficient.

---

## Working with Results

### Extracting fitted models

Set `return_fitted = TRUE` to access the full fitted model objects and their coefficients:

```{r fitted-models}
fit_full <- gformula(
  data        = nonsurvivaldata,
  id_var      = "id",
  time_var    = "time",
  base_vars   = "V",
  exposure    = "A",
  models      = models_bin,
  intervention = list(natural = NULL, always = 1),
  init_recode = init_rc,
  in_recode   = in_rc,
  mc_sample   = 5000,
  R           = 1,
  return_fitted = TRUE,
  quiet       = TRUE,
  seed        = 2025
)

# Names correspond to the response variable of each model
names(fit_full$fitted_models)

# Access a specific model's coefficients
coef(fit_full$fitted_models$A)
```

### Retrieving the simulated data

Set `return_data = TRUE` to retrieve the full Monte Carlo dataset (can be large):

```{r return-data, eval=FALSE}
fit_data <- gformula(..., return_data = TRUE)

# Simulated trajectories with predicted outcomes, one row per MC subject per time
head(fit_data$sim_data)
```

---

## Causal Assumptions

Valid inference requires the following assumptions to hold for the observed data:

1. **Consistency**: the potential outcome under the observed exposure history equals the observed outcome.
2. **Positivity**: every covariate pattern that occurs under the intervention also occurs in the observed data (non-zero probability of receiving each exposure level).
3. **Sequential exchangeability**: no unmeasured confounding of the exposure–outcome relationship at each time point, conditional on the measured past.

For **natural effects** (`mediation_type = "N"`), an additional assumption is required:

4. **No unmeasured exposure-induced mediator–outcome confounding**: there are no confounders of the mediator–outcome relationship that are themselves caused by prior exposure. Interventional effects (`mediation_type = "I"`) do not require this assumption.

---

## Non-standard Covariate Distributions

The four built-in `var_type` values cover the most common cases:

| `var_type` | Distribution |
|---|---|
| `"binary"` | Bernoulli (logistic regression) |
| `"normal"` | Gaussian (linear regression) |
| `"categorical"` | Multinomial (`nnet::multinom`) |
| `"custom"` | User-supplied fit and simulation functions |

For distributions not in this list — **bounded normal**, **truncated normal**, **zero-inflated normal**, **absorbing states** — use `var_type = "custom"` with the `custom_fit` and `custom_sim` arguments to `spec_model()`:

```{r custom-type, eval=FALSE}
# Example: bounded normal covariate in [0, 1]
my_fit <- function(formula, data, ...) {
  # fit any model you like; return the fitted object
  lm(formula, data = data)
}

my_sim <- function(model, newdt, ...) {
  pred <- predict(model, newdata = newdt)
  # draw from N(pred, sigma^2) and clamp to [0, 1]
  draws <- rnorm(length(pred), pred, sigma(model))
  pmin(pmax(draws, 0), 1)
}

m_bounded <- spec_model(
  X ~ A + L + time,
  var_type   = "custom",
  mod_type   = "covariate",
  custom_fit = my_fit,
  custom_sim = my_sim
)
```

`custom_fit(formula, data, ...)` is called during model fitting and must return an object that `custom_sim` knows how to predict from. `custom_sim(model, newdt, ...)` is called at each simulation step and must return a vector of simulated values of length `nrow(newdt)`.

---

## Reference

- Westreich, D., Cole, S. R., Young, J. G., et al. (2012). The parametric g-formula to estimate the effect of highly active antiretroviral therapy on incident AIDS or death. *Statistics in Medicine*, 31, 2000–2009.
- Keil, A. P., Edwards, J. K., Richardson, D. B., Naimi, A. I., & Cole, S. R. (2014). The parametric g-formula for time-to-event data: intuition and a worked example. *Epidemiology*, 25(6), 889–897.
- McGrath, S., Lin, V., Zhang, Z., et al. (2020). gfoRmula: An R package for estimating the effects of sustained treatment strategies via the parametric g-formula. *Patterns*, 1, 100008.
- Zheng, W., & van der Laan, M. (2017). Longitudinal mediation analysis with time-varying mediators and exposures, with application to survival outcomes. *Journal of Causal Inference*, 5(2), 20160006.
- Lin, S.-H., Young, J. G., Logan, R., & VanderWeele, T. J. (2017). Mediation analysis for a survival outcome with time-varying exposures, mediators, and confounders. *Statistics in Medicine*, 36, 4153–4166.
- VanderWeele, T. J., & Tchetgen Tchetgen, E. J. (2017). Mediation analysis with time varying exposures and mediators. *Journal of the Royal Statistical Society: Series B*, 79(3), 917–938.
- Yamamuro, S., Shinozaki, T., Iimuro, S., & Matsuyama, Y. (2021). Mediational g-formula for time-varying treatment and repeated-measured multiple mediators: Application to atorvastatin's effect on cardiovascular disease via cholesterol lowering and anti-inflammatory actions in elderly type 2 diabetics. *Statistical Methods in Medical Research*, 30(8), 1782–1799.
