---
title: "Causal Mediation Analysis with causalMed"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    highlight: null
vignette: >
  %\VignetteIndexEntry{Causal Mediation Analysis with causalMed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Introduction

`mediation()` decomposes the total effect of a time-varying exposure into a
**direct** component (not through the mediator) and an **indirect** component
(through the mediator), for time-varying mediators and confounders, including
confounders affected by prior exposure, the setting the mediational g-formula
was developed for (Lin et al. 2017; VanderWeele & Tchetgen Tchetgen 2017). This
vignette covers:

1. the two estimands (interventional and natural effects) and how to choose,
2. requirements and defaults,
3. a worked single-mediator analysis on a survival outcome, reading every row
   of the output, and the `n_vw` permutation-averaging parameter,
4. multiple mediators, replicating the Yamamuro et al. (2021) simulation,
   with estimates checked against documented true values,
5. censoring,
6. restricting models with `subset` (absorbing states),
7. bootstrap confidence intervals, and
8. natural effects, intermediate confounding, and the targeted (TMLE)
   estimator.

The shared vocabulary (long-format data, `spec_model()`, the `recodes()` lag
hooks) is introduced in `vignette("causalMed-01-overview")`; total-effect
estimation with `gformula()` is in `vignette("causalMed-03-gformula")`.

```{r load}
library(causalMed)
library(data.table)
```

---

## The two estimands

For **natural** effects (`mediation_type = "N"`) the decomposition is exact:

$$\underbrace{E[Y_{1,M(1)}] - E[Y_{0,M(0)}]}_{\text{Total effect}} =
  \underbrace{E[Y_{1,M(0)}] - E[Y_{0,M(0)}]}_{\text{Direct effect}} +
  \underbrace{E[Y_{1,M(1)}] - E[Y_{1,M(0)}]}_{\text{Indirect effect}}$$

where $Y_{a, M(a^*)}$ is the potential outcome under exposure $a$ with the
mediator at its natural value under $a^*$. The mediator is drawn from its
**conditional** distribution: at each time step it is predicted from the
individual's own covariate history, with the exposure argument in the mediator
model swapped to the other level.

For **interventional** effects (`mediation_type = "I"`, the default) the
mediator is instead set to a *stochastic* draw $G_{a^*}$ from its marginal
distribution under $a^*$ (drawn independently across mediators). The direct
and indirect effects are defined relative to these draws and sum to the
**interventional overall effect** $E[Y_{1,G_1}] - E[Y_{0,G_0}]$, which
generally differs from the natural total effect $E[Y_1] - E[Y_0]$.
`mediation()` therefore reports the total effect from a separate
natural-course pass and adds a row `TE - (Direct + Indirect)`, the
decomposition residual ($TE$ minus the overall effect). For natural
effects that residual is exactly zero and the row is omitted.

The key differences:

| | Interventional (`"I"`) | Natural (`"N"`) |
|--|----------------------|-----------------|
| Mediator drawn from | Marginal distribution (permutation) | Conditional distribution (individual history) |
| Cross-world independence required | No | Yes |
| Identified with exposure-affected confounders | Yes | **No** |
| Interpretation | Shift in mediator *population* distribution | Hypothetical individual-level swap |
| Reference | Lin et al. (2017) | Zheng & van der Laan (2017) |

The identifiability row is the practical decision point: natural effects are
**not point-identified** when a confounder of the mediator–outcome
relationship is itself affected by the exposure, an *intermediate confounder*
(Avin, Shpitser & Pearl 2005; VanderWeele & Tchetgen Tchetgen 2017).
`mediation()` warns when a covariate model carries the exposure on its
right-hand side, i.e. when a covariate is *modelled as* exposure-affected. That
is a scan of your formulas, not of the causal structure: it cannot confirm that
such a covariate also confounds the mediator–outcome relationship, and its
silence does not establish that no intermediate confounder exists. Deciding
that is yours. VanderWeele & Tchetgen Tchetgen (2017) propose the interventional
analogues for this setting, which is why `"I"` is the default. The natural
effects section below shows the warning in action.

## Requirements and defaults

**Binary exposure.** The exposure variable must be coded as `0`
(reference/untreated) and `1` (active/treated). Both Lin et al. (2017) and
Zheng & van der Laan (2017) are defined for binary exposures; `mediation()`
stops with an informative error if other values are found.

**One or more mediators.** At least one `mod_type = "mediator"` model is
required. Multiple mediators are supported for `mediation_type = "I"`
(Yamamuro et al. 2021); list them in temporal order. `mediation_type = "N"`
is single-mediator only.

**Temporal ordering.** The list order determines the simulation sequence. Two
orderings are common:

- `A → L → M → S`: confounders at time *t* are not affected by the mediator
  at time *t* (mediator conditions on L; outcome conditions on A, M, L).
- `A → M → L → S`: confounders at time *t* are affected by both exposure and
  mediator (the DAG of Lin et al. 2017).

The function checks that exposure precedes the mediator and that the mediator
precedes the outcome, and warns if either is violated.

**Joint mediator trajectory (interventional effects).** For
`mediation_type = "I"`, a natural-course Monte Carlo cohort is simulated under
each treatment level $a^*$ and every individual's full mediator trajectory
$M(1{:}T)$ is stored in a pool. Each intervention that fixes a
mediator to its $a^*$ value, **including the reference interventions**
$\Phi_{00} = E[Y_{0,G_0}]$ and $\Phi_{11} = E[Y_{1,G_1}]$ and not only the
cross-world $\Phi_{10}$, permutes the relevant pool once and assigns
subject $i$ the *entire* trajectory of one randomly chosen pool individual
(each mediator permuted independently). This is the joint-trajectory algorithm
described by Yamamuro et al. (2021, Figure 3 step 3) and implemented by the
SAS `mGFORMULA` macro (Lin et al. 2017 eAppendix); it samples the marginal
mediator distribution targeted by Lin et al. (2017, Eq. 4) and VanderWeele &
Tchetgen Tchetgen (2017). The pool is not survival-weighted; the full
reference cohort is used at every time step (matching both reference SAS
implementations).

**Warning summary.** Warnings from model fitting (e.g., convergence,
near-separation) are held and printed as a deduplicated summary at function
exit. Repeated warnings (e.g., across 500 bootstrap replicates) are shown once
with a count.

**Outcome types.** `mediation()` handles binary end-of-follow-up outcomes
(`mod_type = "outcome"`) and survival outcomes (`mod_type = "survival"`, with
optional censoring) through the same interface; the outcome model's
`mod_type` is the only difference. This vignette's worked examples use a
survival outcome, the setting the mediational g-formula was developed for
(Lin et al. 2017); a compact binary-outcome example is the mediation quick
start in `vignette("causalMed-01-overview")`.

---

## Survival data

With a survival outcome, `mediation()` (and `gformula()`) model the
**discrete-time hazard**: at each time point, a binary model with
`mod_type = "survival"` estimates the probability of the event among those
still at risk. The simulation accumulates these hazards into a cumulative
incidence, $1 - \prod_t (1 - h_t)$, so all reported quantities, both the
per-intervention estimates and the effect decomposition, are **risks of the
event by the end of follow-up**.

The data must be in long format with one row per subject per period **at
risk**: once the event occurs, no later rows for that subject may be present.

```{r data}
data("survivaldata")
dat <- as.data.table(survivaldata)
head(dat, 8)
```

`survivaldata` contains 3 000 subjects followed over five periods
(`time` = 0–4):

| Variable | Role |
|----------|------|
| `id` | Subject identifier |
| `time` | Time index |
| `V` | Time-fixed baseline covariate |
| `A` | Time-varying binary exposure |
| `L` | Time-varying continuous confounder |
| `M` | Time-varying binary mediator |
| `Y` | Event indicator (1 = event in this period) |
| `C` | Loss to follow-up in this period |
| `lag1_A`, `lag1_L`, `lag1_M` | Previous-period values |

The data-generating ordering within each period is **A → L → M → Y**
(see `?survivaldata`): exposure first, then the confounder (affected by
current exposure), then the mediator (affected by current exposure and
confounder), then the hazard.

---

## Single Mediator: Interventional Effects

Models are listed in the temporal order above. Note `mod_type = "survival"`
for the outcome; the formula may include exposure–mediator interaction terms.

```{r models}
init_s <- recodes(lag1_A = 0, lag1_L = 0, lag1_M = 0)
in_s   <- recodes(lag1_A = A, lag1_L = L, lag1_M = M)

models_surv <- list(
  spec_model(A ~ V + lag1_A + lag1_L + time,
             var_type = "binary", mod_type = "exposure"),
  spec_model(L ~ V + A + lag1_L + time,
             var_type = "normal", mod_type = "covariate"),
  spec_model(M ~ V + A + L + lag1_M + time,
             var_type = "binary", mod_type = "mediator"),
  spec_model(C ~ V + L + time,
             var_type = "binary", mod_type = "censor"),    # loss to follow-up
  spec_model(Y ~ V + A + M + L + A:M + time,
             var_type = "binary", mod_type = "survival")   # discrete-time hazard
)
```

About 9% of subjects are lost to follow-up in `survivaldata`, so a
`mod_type = "censor"` model is declared above. Its role is narrower than it may
appear. Under an intervention the package sets the censoring indicator
to zero and accumulates the hazard over the full follow-up, so the estimand is
the risk that *would* be seen if loss to follow-up were eliminated. Declaring
the censoring model does not, in the g-computation path, alter the
intervention-specific risks: with censoring independent of the counterfactual
event process given the modelled history, the component models are already
fitted without bias on the at-risk rows. The declaration documents the censoring
process and is used by the targeted estimator (`estimator = "tmle"`).

```{r fit-single}
fit_surv <- mediation(
  data           = dat,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y",
  models         = models_surv,
  init_recode    = init_s,
  in_recode      = in_s,
  mediation_type = "I",
  mc_sample      = 20000,
  R              = 1,      # set R > 1 for bootstrap CIs (see below)
  quiet          = TRUE,
  seed           = 2026
)
```

### Reading the per-intervention table

```{r interventions-single}
fit_surv$effect_size
```

Each row is the simulated cumulative incidence under one intervention:

- **`nat0` / `nat1`**: exposure fixed to 0 / 1, mediator following its
  fitted model given each individual's own history. Their contrast is the
  natural plug-in **total effect**.
- **`Phi00` / `Phi11`**: exposure fixed to 0 / 1, mediator drawn from the
  *permuted marginal pool* collected under that same exposure level
  ($E[Y_{0,G_0}]$, $E[Y_{1,G_1}]$). These are the interventional references.
- **`Phi10`**: the cross-world intervention. Exposure fixed to 1, mediator
  drawn from the a = 0 pool ($E[Y_{1,G_0}]$).

### Reading the decomposition

```{r decomp-single}
fit_surv$estimate
```

Row by row:

- **Indirect effect (IIE)** = `Phi11 − Phi10`: the change in risk from
  shifting the *population distribution* of the mediator from its
  never-treated to its always-treated form, while exposure is held at 1.
- **Direct effect (IDE)** = `Phi10 − Phi00`: the effect of exposure with the
  mediator distribution held at its never-treated form.
- **Total effect (TE)** = `nat1 − nat0`: the ordinary g-formula total effect.
- **TE − (Direct + Indirect)**: IDE + IIE sum to the *interventional overall
  effect* `Phi11 − Phi00`, not to TE; this row is the arithmetic difference
  between the two, reported so the decomposition can be read against the total
  effect. Yamamuro et al. (2021) report the corresponding quantity for their
  simulation. (For `mediation_type = "N"` the decomposition is exact and this
  row is absent.)
- **Mediation Proportion** = (TE − IDE) / TE × 100: the share of the total
  effect *not* acting directly, i.e. the indirect effects plus the residual
  (Yamamuro et al. 2021).
- **Mediation Proportion (multiplicative)** =
  $RR_{IDE}(RR_{IIE}-1)/(RR_{OE}-1) \times 100$ on the risk-ratio scale
  (Lin et al. 2017, Table 2).

The `RD` column is the risk-difference scale, `RR` the risk-ratio scale
(`RR` is not applicable to the residual and proportion rows). With `R > 1`
both scales get bootstrap standard errors and percentile/normal confidence
intervals.

`print(fit_surv)` displays both tables together with the intervention
definitions as a legend, an analysis-setup summary (mediator, outcome,
data dimensions, `n_vw`), and an observed (nonparametric) cumulative-incidence
benchmark computed directly from the data.

### The `n_vw` argument

Every intervention that draws its mediators from a permuted pool does so by
randomly permuting the pool of simulated mediator trajectories. Under
`mediation_type = "I"` that is the *whole* decomposition (the references
$\Phi_{00}$ and $\Phi_{11}$ as well as the cross-world $\Phi_{10}$), but not
the natural-course interventions `nat0`/`nat1`, whose mediators come from their own
fitted models. `n_vw` controls how many independent permutations are averaged
per intervention (default `2`, matching the SAS `mGFORMULA` macro). Averaging
reduces Monte Carlo noise from the permutation step at the cost of one extra
simulation pass per intervention; `n_vw = 1` is faster and slightly noisier. It
has no effect on natural effects (`mediation_type = "N"`), which do not use
permutation.

One consequence worth knowing if you use `return_data = TRUE`: with `n_vw > 1`,
`sim_data` keeps only the last permutation of each pool-drawing intervention while
`effect_size$Est` averages all of them, so recomputing `mean(Pred_Y)` from
`sim_data` will not exactly reproduce `Est` for those interventions. Set
`n_vw = 1` if you need the two to agree.

```{r nvw, eval=FALSE}
mediation(..., n_vw = 1)   # single permutation per intervention: faster, noisier
```

---

## Multiple Mediators: the Yamamuro et al. (2021) Simulation

With `mediation_type = "I"`, any number of mediators can be analysed by
supplying several `mod_type = "mediator"` models **in temporal order**. We
demonstrate this on `yamamurodata`, a dataset simulated from the
data-generating process of the Yamamuro et al. (2021) simulation study:
a time-varying treatment `A`, a confounder `L`, two sequential mediators
`M1` and `M2`, and a survival outcome over three visits, ordered
**A → L → M1 → M2 → Y** within each visit. Because the process is known,
the **true interventional effects are documented in `?yamamurodata`**
(computed at $n = 10^7$), so the estimates below can be checked against
the truth.

```{r yam-data}
data("yamamurodata")
yam <- as.data.table(yamamurodata)
head(yam, 6)
```

Three specification details matter here:

- **Visit indicators.** The published models give each visit its own
  intercept. Write these as `I(as.integer(time == k))` rather than
  `factor(time)`: during simulation each Monte Carlo time slice carries a
  single `time` value, so a factor would drop the unobserved levels while
  the numeric indicator predicts safely at every step.
- **`subset = time > 0`.** The visit-0 values of `A`, `L`, `M1`, `M2` are
  baseline draws, not model output, so the time-varying models are fitted
  and simulated only for `time > 0`. Visit 0 is seeded in the Monte Carlo
  cohort from the time-fixed baseline columns (`L0base`, `M10base`,
  `M20base`) via `init_recode`.
- The models include the quadratic and lag terms of the published
  "correctly specified" scenario.

```{r yam-models}
models_yam <- list(
  spec_model(A ~ V + lag1_A + lag1_L + I(lag1_L^2) + lag1_M1 + lag1_M2 +
               I(as.integer(time == 2)),
             var_type = "binary", mod_type = "exposure", subset = time > 0),
  spec_model(L ~ V + A + lag1_L + I(lag1_L^2) + lag1_M1 + lag1_M2 +
               I(as.integer(time == 2)),
             var_type = "normal", mod_type = "covariate", subset = time > 0),
  spec_model(M1 ~ V + A + L + I(L^2) + lag1_L + I(lag1_L^2) + lag1_M1 + lag1_M2 +
               I(as.integer(time == 2)),
             var_type = "normal", mod_type = "mediator", subset = time > 0),
  spec_model(M2 ~ V + A + L + I(L^2) + lag1_L + I(lag1_L^2) + lag1_M1 + lag1_M2 +
               I(as.integer(time == 2)),
             var_type = "normal", mod_type = "mediator", subset = time > 0),
  spec_model(Y ~ V + A + L + I(L^2) + M1 + M2 + M1:M2 +
               I(as.integer(time == 1)) + I(as.integer(time == 2)),
             var_type = "binary", mod_type = "survival")
)

fit_yam <- mediation(
  data           = yam,
  id_var         = "id",
  time_var       = "time",
  base_vars      = c("V", "L0base", "M10base", "M20base"),
  exposure       = "A",
  outcome        = "Y",
  models         = models_yam,
  init_recode    = recodes(L = L0base, M1 = M10base, M2 = M20base,
                           lag1_A = 0, lag1_L = 0, lag1_M1 = 0, lag1_M2 = 0),
  in_recode      = recodes(lag1_A = A, lag1_L = L, lag1_M1 = M1, lag1_M2 = M2),
  mediation_type = "I",
  n_vw           = 2,      # matches the SAS mGFORMULA macro
  mc_sample      = 20000,
  R              = 1,
  quiet          = TRUE,
  seed           = 2026
)

fit_yam$effect_size
```

For $N$ mediators the intervention list grows to $4 + N$: the intermediate
interventions `Phi1_k` switch the first $k$ mediators to the a = 1 pool while the rest
stay at the a = 0 pool. Each mediator's indirect effect is the **sequential
contrast** $IIE(M_k) = \Phi_{1,k} - \Phi_{1,k-1}$, labelled
`Indirect effect (<name>)` in the output, and
$IDE + \sum_k IIE(M_k)$ equals the interventional overall effect
`Phi11 − Phi00`. Each mediator's pool is permuted independently.

### Comparing against the true values

```{r yam-truth}
truth <- data.table(
  Effect = c("Total effect", "Direct effect", "Indirect effect (M1)",
             "Indirect effect (M2)", "TE - (Direct + Indirect)"),
  True   = c(-6.36, -3.20, -2.29, -0.97, 0.10)   # from ?yamamurodata
)
est <- as.data.table(fit_yam$estimate)[, .(Effect, Estimate = RD * 100)]
merge(truth, est, by = "Effect", sort = FALSE)
```

The estimates reproduce the structure of the truth: all signs and the relative
magnitudes (IDE > IIE via M1 > IIE via M2, small positive residual) are
recovered. The two indirect effects land within 0.05 percentage points of their
true values, while the total and direct effects are both about 1.4 percentage
points further from zero than the truth. The `True` column is a large-sample
value computed from the data-generating process, whereas the estimate comes from
one finite dataset simulated a finite number of times, so it carries both
sampling error and Monte Carlo error; the run above is a single point estimate
with neither quantified. A bootstrap (`R > 1`) is what supplies that uncertainty
in applied use.

Because the decomposition is sequential, the *order* of the mediator models
matters: it must reflect the assumed temporal/causal ordering among the
mediators. Multiple mediators are not available for `mediation_type = "N"`
(the natural-effects references define a single mediator only); `mediation()`
stops with an error in that case.

---

## Censoring

If follow-up can end for reasons other than the event, include a censoring
indicator with `mod_type = "censor"`. To illustrate, we censor some
follow-up in `survivaldata` with probability depending on the confounder
(informative censoring), ending each subject's follow-up at their first
censoring:

```{r censor-data}
set.seed(11)
dat_c <- copy(dat)
dat_c[, C := rbinom(.N, 1, plogis(-3.5 + 0.8 * L))]
dat_c[, after_cens := cumsum(shift(C, fill = 0)), by = id]
dat_c <- dat_c[after_cens == 0][, after_cens := NULL]
dat_c[C == 1, Y := 0L]   # censored before the event in that period
```

Add a censoring model to the list (after the covariates/mediator it depends
on, before the survival model):

```{r fit-censor}
models_cens <- list(
  spec_model(A ~ V + lag1_A + lag1_L + time,
             var_type = "binary", mod_type = "exposure"),
  spec_model(L ~ V + A + lag1_L + time,
             var_type = "normal", mod_type = "covariate"),
  spec_model(M ~ V + A + L + lag1_M + time,
             var_type = "binary", mod_type = "mediator"),
  spec_model(C ~ V + A + L + time,
             var_type = "binary", mod_type = "censor"),     # censoring model
  spec_model(Y ~ V + A + M + L + A:M + time,
             var_type = "binary", mod_type = "survival")
)

fit_cens <- mediation(
  data           = dat_c,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y",
  models         = models_cens,
  init_recode    = init_s,
  in_recode      = in_s,
  mediation_type = "I",
  mc_sample      = 20000,
  R              = 1,
  quiet          = TRUE,
  seed           = 2026
)

fit_cens$estimate
```

In the simulation, censoring is **abolished** in every intervention (all
interventions fix the exposure, and the censoring indicator is set to 0), so
the reported risks are counterfactual risks *in the absence of censoring*, as
in the g-formula treatment of right-censoring described by Robins (1986) and
Westreich et al. (2012). The censoring model's role is to let the hazard model
be fitted on data where censoring depends on measured covariates. Whether these
estimates recover the full-data ones depends on censoring being independent of
the outcome given the modelled covariates, an assumption about your data, not
something the package can check. Compare the two tables here to see how they
came out in this constructed example.

---

## Restricting Models with `subset` (Absorbing States)

`spec_model(subset = ...)` fits and simulates a model only on rows meeting a
condition. The classic use is an **absorbing state**. In the GvHD analysis of
Keil et al. (2014), for instance, the exposure can only switch on once, so its
model is estimated among the not-yet-exposed and the value is carried forward
deterministically afterwards:

```{r subset, eval=FALSE}
# exposure can occur only while gvhdm1 == 0 …
spec_model(gvhd ~ all + cmv + male + age + ...,
           var_type = "binary", mod_type = "exposure",
           subset = gvhdm1 == 0)

# … and is locked at 1 afterwards via the end-of-step hook
out_recode = recodes(gvhd = ifelse(gvhdm1 == 1, 1, gvhd))
```

Rows excluded by `subset` keep their current value at that step, so the
`out_recode` carry-forward completes the absorbing behaviour. The package
ships the `gvhd` dataset (see `?gvhd`) used in that paper, and a complete
total-effect g-formula analysis on it (three absorbing states, a censoring
model, and restricted cubic splines) is walked through in
`vignette("causalMed-03-gformula")` (Keil et al. 2014 describe the analysis in
full, including the spline knots).

---

## Bootstrap Confidence Intervals

As elsewhere in the package, set `R > 1` for subject-level bootstrap CIs and
optionally register a parallel plan first (the general bootstrap mechanics,
percentile versus normal intervals and parallel plans, are covered in
`vignette("causalMed-03-gformula")`):

```{r boot, eval=FALSE}
library(future)
plan(multisession)

fit_ci <- mediation(
  data           = dat,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y",
  models         = models_surv,
  init_recode    = init_s,
  in_recode      = in_s,
  mediation_type = "I",
  mc_sample      = 20000,
  R              = 500,
  seed           = 2026
)

plan(sequential)

# estimate now carries Sd, percentile and normal CIs on the RD and RR scales
fit_ci$estimate

# the per-replicate bootstrap draws are also retained, for custom diagnostics
# (e.g. counting non-finite Mediation Proportion replicates):
fit_ci$boot_estimates$effects
```

---

## Natural Effects and Intermediate Confounding

`mediation_type = "N"` (Zheng & van der Laan 2017) is also defined for
survival outcomes. However, in this dataset the confounder `L` is affected by
the current exposure, an *intermediate confounder*, and natural direct and
indirect effects are **not identifiable** in that setting (see the estimands
section above). `mediation()` finds the exposure on the right-hand side of a
covariate model and warns:

```{r natural, warning=TRUE}
fit_nat <- mediation(
  data           = dat,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y",
  models         = models_surv,
  init_recode    = init_s,
  in_recode      = in_s,
  mediation_type = "N",
  mc_sample      = 20000,
  R              = 1,
  quiet          = TRUE,
  seed           = 2026
)
```

Interventional effects (`"I"`) remain identifiable under intermediate
confounding, and VanderWeele & Tchetgen Tchetgen (2017) propose them for
exactly this setting; that is why `"I"` is the package default. The choice
between the two estimands is a substantive one; see Miles (2023) on what a
non-zero interventional indirect effect does and does not establish.

### A targeted (TMLE) estimator for natural effects

When the natural-effects estimand *is* appropriate for your data, the plug-in
simulation is not the only estimator: `estimator = "tmle"` implements the
targeted maximum likelihood estimator from the same reference (Zheng & van der
Laan 2017, Section 4.3), including right-censored survival outcomes. Instead
of simulating forward, it runs backward iterated regressions with targeted
fluctuation steps, weighting by inverse treatment *and* censoring
probabilities. Zheng & van der Laan (2017) establish multiple robustness for
it, meaning consistency when certain subsets of the nuisance models are correct,
and it reports Wald CIs from the efficient influence curve, so no bootstrap is
required (`R` and `mc_sample` are ignored). Those results are stated for
correctly specified nuisance models; this implementation builds its targeted
sequential regressions as additive main-effects working models, so
transformations and interactions written into your formulas are not carried
into them (see `?mediation`).

Reusing the censored dataset and model list from the censoring section (which
already includes the required exposure model):

```{r tmle-survival}
fit_tmle_s <- mediation(
  data           = dat_c,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y",
  models         = models_cens,
  init_recode    = init_s,
  in_recode      = in_s,
  mediation_type = "N",
  estimator      = "tmle",
  quiet          = TRUE,
  seed           = 2026
)

fit_tmle_s$estimate
```

Note the **recode restriction**: the targeted engine evaluates only lag-style
recodes, so `in_recode` entries must copy a single column (as `init_s`/`in_s` do
here), lags of the exposure must copy the exposure itself (chained lags like
`lag2_A = lag_A` are rejected), `init_recode` entries must be a constant or a
column name, and
`out_recode` is not supported. Derived recodes (splines, cumulative counts,
carry-forward flags such as the absorbing-state pattern shown earlier) and
deeper exposure history require `estimator = "gcomp"`. Violations raise an error rather than being silently
dropped, so a distorted TMLE cannot reach you unnoticed.

Two caveats:

- **The identification warning still applies.** The TMLE changes the
  *estimator*, not the estimand: if an intermediate confounder makes natural
  effects unidentifiable (as in this dataset, where `L` depends on current
  `A`), no estimator can fix that. The same warning is emitted at run time
  and re-surfaced as a short caveat under the decomposition when you
  `print()` the result (the offending confounders are also stored in
  `fit$intermediate_confounders`). It is for this setting that VanderWeele &
  Tchetgen Tchetgen (2017) propose the interventional estimand.
- **Positivity warnings deserve attention.** If few subjects follow an
  intervened regime (e.g., never treated at every time point), the affected
  targeting steps are skipped with a collected warning and the corresponding
  functional leans on model extrapolation, exactly as the plug-in does, but
  with the sparse support made visible instead of silent. Neither estimator
  can conjure information the data do not contain; the TMLE tells you when
  that is happening, where the plug-in produces a seemingly precise answer by
  extrapolating from its parametric models. When the two estimators disagree
  materially, the first things to inspect are these positivity warnings and the
  model specifications. Relatedly, the influence-curve Wald CIs can be
  anti-conservative for functionals under positivity strain: in a replication of
  the reference paper's heavily censored simulation, coverage for the never-treat
  functionals was about 87% rather than the nominal 95% even at n = 4000, while
  the treated functionals covered at nominal rates. Interpret CIs for sparsely
  supported regimes with corresponding caution.

The per-subject influence-curve values are returned in
`fit_tmle_s$tmle_diag$eic` (one column per functional) for custom contrasts
or diagnostics; `fit_tmle_s$tmle_diag$eic_mean` should be near zero for
well-supported functionals.

In simulations replicating the survival design of Zheng & van der Laan
(2017, Section 5), this implementation recovers that paper's reported value for
the cross-world functional, $J^{1,0} \approx 0.912$. Its sequential regressions
are GLMs rather than the Super Learner fits used there, so the efficiency
results reported in the paper do not carry over to it unchanged.

---

## References

- Lin, S.-H., Young, J. G., Logan, R., & VanderWeele, T. J. (2017). Mediation
  analysis for a survival outcome with time-varying exposures, mediators, and
  confounders. *Statistics in Medicine*, 36, 4153–4166.
- VanderWeele, T. J., & Tchetgen Tchetgen, E. J. (2017). Mediation analysis
  with time varying exposures and mediators. *Journal of the Royal
  Statistical Society: Series B*, 79(3), 917–938.
- Yamamuro, S., Shinozaki, T., Iimuro, S., & Matsuyama, Y. (2021).
  Mediational g-formula for time-varying treatment and repeated-measured
  multiple mediators. *Statistical Methods in Medical Research*, 30(8),
  1782–1799.
- Zheng, W., & van der Laan, M. (2017). Longitudinal mediation analysis with
  time-varying mediators and exposures, with application to survival
  outcomes. *Journal of Causal Inference*, 5(2).
- Avin, C., Shpitser, I., & Pearl, J. (2005). Identifiability of path-specific
  effects. *Proceedings of the 19th International Joint Conference on
  Artificial Intelligence (IJCAI)*, 357–363.
- Miles, C. H. (2023). On the causal interpretation of randomised
  interventional indirect effects. *Journal of the Royal Statistical Society:
  Series B*, 85(4), 1154–1172.
- Keil, A. P., Edwards, J. K., Richardson, D. B., Naimi, A. I., & Cole, S. R.
  (2014). The parametric g-formula for time-to-event data: intuition and a
  worked example. *Epidemiology*, 25(6), 889–897.
- Westreich, D., Cole, S. R., Young, J. G., et al. (2012). The parametric
  g-formula to estimate the effect of highly active antiretroviral therapy on
  incident AIDS or death. *Statistics in Medicine*, 31, 2000–2009.
