---
title: "Getting Started with causalMed"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    highlight: null
vignette: >
  %\VignetteIndexEntry{Getting Started with causalMed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Introduction

**causalMed** implements the parametric g-formula for two related goals:

1. **Total effect estimation** (`gformula()`): estimate the counterfactual
   mean outcome had everyone in the study followed a given exposure strategy,
   using the standard parametric g-formula of Westreich et al. (2012) and
   McGrath et al. (2020).

2. **Causal mediation analysis** (`mediation()`): decompose the total effect
   into a direct component (not through the mediator) and an indirect
   component (through the mediator), using either **interventional**
   direct/indirect effects (Lin et al. 2017) or **natural** direct/indirect
   effects (Zheng & van der Laan 2017).

Both handle time-varying exposures, mediators, and confounders — including
confounders that are themselves affected by prior exposure. This is the setting
the g-formula was introduced for (Robins 1986), and the one in which natural
direct and indirect effects are not identifiable (Avin, Shpitser & Pearl 2005;
VanderWeele & Tchetgen Tchetgen 2017).

This vignette is the short tour: the data format, the model-specification
vocabulary that every analysis uses, and one quick-start example of each
function. The full contents live in their own vignettes:

| Vignette | Contents |
|---|---|
| `vignette("causalMed-02-mediation")` | Mediation: estimands, reading the decomposition, multiple mediators, censoring, natural effects, and the targeted (TMLE) estimator. |
| `vignette("causalMed-03-gformula")` | Total effects: binary and survival outcomes, dynamic interventions, a published replication (Keil et al. 2014), bootstrap CIs, results handling, custom distributions. |
| `vignette("causalMed-04-vs-gfoRmula")` | How the total-effect engine compares with the CRAN reference implementation `gfoRmula`. |

```{r load}
library(causalMed)
library(data.table)
```

---

## Data Structure

The package requires **long-format** data: one row per subject per time point.

```{r data}
data("nonsurvivaldata")
head(nonsurvivaldata, 10)
```

The `nonsurvivaldata` dataset contains 3 000 subjects observed at five time
points (0, 1, 2, 3, 4), 15 000 rows in total:

| Variable | Role |
|----------|------|
| `id` | Subject identifier |
| `time` | Time index (0, 1, 2, 3, 4) |
| `V` | Time-fixed baseline covariate |
| `A` | Time-varying binary exposure |
| `L1` | Time-varying continuous confounder |
| `L2` | Time-varying binary confounder |
| `M` | Time-varying continuous mediator |
| `Y_bin` | Binary outcome (end-of-follow-up) |

The temporal ordering within each period is **A → L → M → Y**: the exposure is
decided first, the confounders respond to it, then the mediator, then the
outcome (this matches the data-generating process documented in
`?nonsurvivaldata`).

---

## Specifying Models

Every variable that is not time-fixed and not the identifier needs a
parametric model. Models are created with `spec_model()` and collected into a
list **in the temporal order they should be simulated**.

```{r spec-model-args, eval=FALSE}
spec_model(
  formula,            # Standard R formula: response ~ predictors
  var_type,           # Distribution for simulation
  mod_type,           # Role in the causal structure
  subset  = NULL,     # Optional condition restricting which rows are used
  recode  = NULL      # Optional within-loop recoding (see recodes())
)
```

### `var_type` — how to draw simulated values

The four built-in `var_type` values cover the most common cases:

| `var_type` | Distribution |
|---|---|
| `"binary"` | Bernoulli (logistic regression) |
| `"normal"` | Gaussian (linear regression), clipped to the observed range by default |
| `"categorical"` | Multinomial (`nnet::multinom`) |
| `"custom"` | User-supplied fit and simulation functions |


**Simulated numeric values are clipped to the observed range of the response by
default.** `gfoRmula` applies the same rule by default, through its `sim_trunc`
argument. Pass `spec_model(truncate = FALSE)` to draw from the untruncated
fitted distribution instead (this also stops the clipping being applied to
`custom_sim` output). Which of the two is appropriate depends on the variable
being simulated.

For distributions still not covered — **truncated normal**, **zero-inflated
normal**, **absorbing states** — use `var_type = "custom"` with the
`custom_fit` and `custom_sim` arguments to `spec_model()`. The example below
implements a **zero-inflated normal** (a point mass at zero, Gaussian
otherwise — e.g. a biomarker that is exactly zero for part of the population)
as a two-part model. The object returned by `custom_fit` can be *any*
structure — here a plain list holding two fits — as long as `custom_sim`
knows how to use it:

```{r custom-type, eval=FALSE}
# Part 1: is the response positive?  Part 2: its level, among the positives.
zin_fit <- function(formula, data, ...) {
  y <- model.response(model.frame(formula, data = data))
  fit_any <- glm(update(formula, I(. > 0) ~ .), family = binomial(), data = data)
  fit_pos <- lm(formula, data = data[y > 0, ])
  list(any = fit_any, pos = fit_pos)
}

zin_sim <- function(model, newdt, ...) {
  p_any  <- predict(model$any, newdata = newdt, type = "response")
  m_pos  <- predict(model$pos, newdata = newdt)
  is_pos <- rbinom(nrow(newdt), 1, p_any)
  pos    <- pmax(rnorm(nrow(newdt), m_pos, sigma(model$pos)), 0)
  is_pos * pos
}

m_zin <- spec_model(
  X ~ A + L + time,
  var_type   = "custom",
  mod_type   = "covariate",
  custom_fit = zin_fit,
  custom_sim = zin_sim,
  truncate   = FALSE   # zin_sim is authoritative over its own range
)
```

`custom_fit(formula, data, ...)` is called once during model fitting and must
return an object that `custom_sim` knows how to predict from.
`custom_sim(model, newdt, ...)` is called at each simulation step and must
return a vector of simulated values of length `nrow(newdt)`. (A plain bounded
normal needs no custom type at all — that is `var_type = "normal"`'s default
clipping behaviour, as described above)


### `mod_type` — causal role

| Value | Description |
|-------|-------------|
| `"covariate"` | Time-varying confounder |
| `"exposure"` | Intervention variable |
| `"mediator"` | Mediator (required for `mediation()`) |
| `"outcome"` | Binary end-of-follow-up outcome |
| `"survival"` | Discrete-time event indicator (hazard model) |
| `"censor"` | Right-censoring indicator |

### Temporal ordering is critical

The list order determines the simulation sequence at each time step, and it
must follow your assumed data-generating mechanism. Confounders at time *t*
that are affected by the exposure at time *t* must come **after** the exposure
model and **before** the outcome model.

---

## Managing Lagged Variables

Most models condition on previous-period values (lags). These are created
with `recodes()`, which captures expressions to be evaluated inside the
simulated dataset at each step.

| Hook | When applied | Typical use |
|------|-------------|-------------|
| `init_recode` | Once at the first time point | Set lag variables to their baseline values |
| `in_recode` | Start of each subsequent time step | Update lags from the previous step's values |
| `out_recode` | End of each time step | Post-simulation transforms (cumulative sums, etc.) |

```{r recodes-example}
init_rc <- recodes(lag1_A  = 0,   # At t=0, all lags initialised to 0
                   lag1_L1 = 0,
                   lag1_L2 = 0)

in_rc   <- recodes(lag1_A  = A,   # At each subsequent step, copy current values
                   lag1_L1 = L1,
                   lag1_L2 = L2)
```

The hooks are required even when the lag columns already exist in your data.
The Monte Carlo cohort is built from `id_var` and `base_vars` only, so any
column a model references that is neither of those — every lag column included —
has to be created by `init_recode` at the first time point and maintained by
`in_recode` thereafter. Without them the run stops with
`object 'lag1_A' not found`. Listing lag columns in `base_vars` is not a
substitute: they are not time-fixed, and `gformula()` warns that doing so
distorts the sampled cohort.


---


## Parallel bootstrap

The bootstrap loop uses `future.apply::future_lapply` internally. Enable
parallelism by setting a parallel plan before calling the function:

```{r parallel, eval=FALSE}
library(future)
plan(multisession)   # use multiple R sessions in parallel

fit_par <- gformula(..., R = 500)

plan(sequential)     # restore default after use
```

On Unix/macOS, `plan(multicore)` forks the current session instead of launching
new ones; see `?future::plan` for the trade-offs.


---

## Quick start: a total effect with `gformula()`

Models in temporal order (**A → L1 → L2 → Y**, matching the documented
data-generating process — exposure first, current `A` in the confounder
models), then three strategies: the natural course, always treat, never treat.

```{r quickstart-gformula}
m_A  <- spec_model(A     ~ V + lag1_A + lag1_L1 + lag1_L2 + time,
                   var_type = "binary",  mod_type = "exposure")
m_L1 <- spec_model(L1    ~ V + A + lag1_L1 + time,
                   var_type = "normal",  mod_type = "covariate")
m_L2 <- spec_model(L2    ~ V + A + lag1_L2 + time,
                   var_type = "binary",  mod_type = "covariate")
m_Y  <- spec_model(Y_bin ~ V + A + L1 + L2,
                   var_type = "binary",  mod_type = "outcome")

fit_bin <- gformula(
  data        = nonsurvivaldata,
  id_var      = "id",
  time_var    = "time",
  base_vars   = "V",
  exposure    = "A",
  models      = list(m_A, m_L1, m_L2, m_Y),
  intervention = list(natural = NULL, always_treat = 1, never_treat = 0),
  ref_int     = "natural",
  init_recode = init_rc,
  in_recode   = in_rc,
  mc_sample   = 10000,
  R           = 1,       # set R > 1 for bootstrap confidence intervals
  quiet       = TRUE,
  seed        = 2025
)

fit_bin$effect_size   # mean outcome under each strategy
fit_bin$estimate      # risk difference / ratio vs the natural course
```

That is the whole workflow: models, interventions, one call. Survival
outcomes, dynamic (rule-based) interventions, the Keil et al. (2014) GvHD
replication, bootstrap confidence intervals, extracting fitted models and
simulated data, and custom covariate distributions are all in
`vignette("causalMed-03-gformula")`.

---

## Quick start: mediation with `mediation()`

Add a mediator model (`mod_type = "mediator"`) and name the outcome. The
default estimand is the **interventional** direct/indirect decomposition
(Lin et al. 2017), which stays identifiable when confounders are affected by
the exposure:

```{r quickstart-mediation}
init_med <- recodes(lag1_A = 0, lag1_L1 = 0, lag1_L2 = 0, lag1_M = 0)
in_med   <- recodes(lag1_A = A, lag1_L1 = L1, lag1_L2 = L2, lag1_M = M)

models_med <- list(
  spec_model(A   ~ V + lag1_L1 + lag1_L2 + lag1_A + time,
             var_type = "binary",  mod_type = "exposure"),
  spec_model(L1  ~ V + A + lag1_L1 + time,
             var_type = "normal",  mod_type = "covariate"),
  spec_model(L2  ~ V + A + lag1_L2 + time,
             var_type = "binary",  mod_type = "covariate"),
  spec_model(M   ~ V + A + L1 + L2 + lag1_M + time,
             var_type = "normal",  mod_type = "mediator"),   # <-- mediator
  spec_model(Y_bin ~ V + A + M + L1 + L2,
             var_type = "binary",  mod_type = "outcome")
)

fit_ide <- mediation(
  data           = nonsurvivaldata,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y_bin",
  models         = models_med,
  init_recode    = init_med,
  in_recode      = in_med,
  mediation_type = "I",     # interventional IDE/IIE (the default)
  mc_sample      = 10000,
  R              = 1,
  quiet          = TRUE,
  seed           = 2025
)

fit_ide$estimate
```

The `estimate` table is the decomposition: the direct effect, the indirect
effect (through `M`), the total effect, a mediated-interaction residual row,
and the proportion mediated. A second estimand — **natural** direct/indirect
effects (`mediation_type = "N"`), with an optional targeted maximum likelihood
estimator (`estimator = "tmle"`) — needs stronger assumptions and is not
identified when a confounder responds to the exposure (as `L1`/`L2` do here).

What the decomposition rows mean, the choice between the two estimands,
survival outcomes, multiple mediators, censoring, and the TMLE are all in
`vignette("causalMed-02-mediation")`.

---

## Causal Assumptions

The parametric g-formula identifies these effects under the following
assumptions (Robins 1986; Westreich et al. 2012; Keil et al. 2014), which the
package cannot check and which hold or fail as a property of your data and
design, not of the code:

1. **Consistency**: the potential outcome under the observed exposure history
   equals the observed outcome.
2. **Positivity**: every covariate pattern that occurs under the intervention
   also occurs in the observed data (non-zero probability of receiving each
   exposure level).
3. **Sequential exchangeability**: no unmeasured confounding of the
   exposure–outcome relationship at each time point, conditional on the
   measured past.

For **natural effects** (`mediation_type = "N"`), an additional assumption is
required:

4. **No unmeasured exposure-induced mediator–outcome confounding**: there are
   no confounders of the mediator–outcome relationship that are themselves
   caused by prior exposure. Natural direct and indirect effects are not
   identifiable when one exists (Avin, Shpitser & Pearl 2005; VanderWeele &
   Tchetgen Tchetgen 2017); the randomized interventional analogues targeted by
   `mediation_type = "I"` are identifiable without this assumption
   (VanderWeele & Tchetgen Tchetgen 2017).

---

## References

- Robins, J. M. (1986). A new approach to causal inference in mortality studies
  with a sustained exposure period — application to control of the healthy
  worker survivor effect. *Mathematical Modelling*, 7(9–12), 1393–1512.
- Westreich, D., Cole, S. R., Young, J. G., et al. (2012). The parametric
  g-formula to estimate the effect of highly active antiretroviral therapy on
  incident AIDS or death. *Statistics in Medicine*, 31, 2000–2009.
- Avin, C., Shpitser, I., & Pearl, J. (2005). Identifiability of path-specific
  effects. *Proceedings of the 19th International Joint Conference on
  Artificial Intelligence (IJCAI)*, 357–363.
- VanderWeele, T. J., & Tchetgen Tchetgen, E. J. (2017). Mediation analysis
  with time varying exposures and mediators. *Journal of the Royal Statistical
  Society: Series B*, 79(3), 917–938.
- McGrath, S., Lin, V., Zhang, Z., et al. (2020). gfoRmula: An R package for
  estimating the effects of sustained treatment strategies via the parametric
  g-formula. *Patterns*, 1, 100008.
- Lin, S.-H., Young, J. G., Logan, R., & VanderWeele, T. J. (2017). Mediation
  analysis for a survival outcome with time-varying exposures, mediators, and
  confounders. *Statistics in Medicine*, 36, 4153–4166.
- Zheng, W., & van der Laan, M. (2017). Longitudinal mediation analysis with
  time-varying mediators and exposures, with application to survival outcomes.
  *Journal of Causal Inference*, 5(2), 20160006.
- Keil, A. P., Edwards, J. K., Richardson, D. B., Naimi, A. I., & Cole, S. R.
  (2014). The parametric g-formula for time-to-event data: intuition and a
  worked example. *Epidemiology*, 25(6), 889–897.
