---
title: "Getting Started with causalMed"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    highlight: null
vignette: >
  %\VignetteIndexEntry{Getting Started with causalMed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Introduction

**causalMed** implements the parametric g-formula for two related goals:

1. **Total effect estimation** (`gformula`): estimate the counterfactual mean outcome had everyone in the study followed a given exposure strategy, using the standard parametric g-formula of Westreich et al. (2012) and McGrath et al. (2020).

2. **Causal mediation analysis** (`mediation`): decompose the total effect into a direct component (not through the mediator) and an indirect component (through the mediator), using either:
   - **Interventional direct/indirect effects (IDE/IIE)** via the survival mediational g-formula of Lin et al. (2017);
   - **Natural direct/indirect effects (NDE/NIE)** via the longitudinal identification formula of Zheng & van der Laan (2017).

Both approaches handle time-varying exposures, mediators, and confounders — including confounders that are themselves affected by prior exposure (exposure-induced mediator–outcome confounding), a setting where standard regression-based mediation fails.

```{r load}
library(causalMed)
library(data.table)
```

---

## Data Structure

The package requires **long-format** data: one row per subject per time point.

```{r data}
data("nonsurvivaldata")
head(nonsurvivaldata, 10)
```

The `nonsurvivaldata` dataset contains 1 000 subjects observed at five time points (0, 1, 2, 3, 4):

| Variable | Role |
|----------|------|
| `id` | Subject identifier |
| `time` | Time index (0, 1, 2, 3, 4) |
| `V` | Time-fixed baseline covariate |
| `A` | Time-varying binary exposure |
| `L1` | Time-varying continuous confounder |
| `L2` | Time-varying binary confounder |
| `M` | Time-varying continuous mediator |
| `Y_bin` | Binary outcome (end-of-follow-up) |

The assumed temporal ordering within each period is **A → M → L → Y**, meaning that within any time point, exposure precedes the mediator, which precedes the confounders, which precede the outcome.

---

## Specifying Models

Every variable that is not time-fixed and not the identifier needs a parametric model. Models are created with `spec_model()` and collected into a list **in the temporal order they should be simulated**.

```{r spec-model-args, eval=FALSE}
spec_model(
  formula,            # Standard R formula: response ~ predictors
  var_type,           # Distribution for simulation
  mod_type,           # Role in the causal structure
  subset  = NULL,     # Optional condition restricting which rows are used
  recode  = NULL      # Optional within-loop recoding (see recodes())
)
```

### `var_type` — how to draw simulated values

| Value | Distribution | Typical use |
|-------|-------------|-------------|
| `"binary"` | logistic regression | Binary covariates, exposure, outcome |
| `"normal"` | linear regression | Continuous covariates, mediator |
| `"categorical"` | Multinomial via `nnet::multinom` | Unordered categorical covariates |
| `"custom"` | User-supplied function | Specialised distributions |

### `mod_type` — causal role

| Value | Description |
|-------|-------------|
| `"covariate"` | Time-varying confounder |
| `"exposure"` | Intervention variable |
| `"mediator"` | Mediator (required for `mediation()`) |
| `"outcome"` | Binary end-of-follow-up outcome |
| `"survival"` | Discrete-time event indicator (hazard model) |
| `"censor"` | Right-censoring indicator |

### Temporal ordering is critical

The list order determines the simulation sequence at each time step. This should follow the data-generating mechanism. Confounders at time *t* that are affected by exposure at time *t* must come **after** the exposure model but **before** the outcome model.

---

## Managing Lagged Variables

Most models condition on previous-time values (lags). These are created using `recodes()`, which captures expressions to be evaluated inside the simulated dataset at each step.

Three hooks are available:

| Hook | When applied | Typical use |
|------|-------------|-------------|
| `init_recode` | Once at the first time point | Set lag variables to their baseline values |
| `in_recode` | Start of each subsequent time step | Update lags from the previous step's values |
| `out_recode` | End of each time step | Post-simulation transforms (cumulative sums, etc.) |

```{r recodes-example}
init_rc <- recodes(lag1_A  = 0,   # At t=0, all lags initialised to 0
                   lag1_L1 = 0,
                   lag1_L2 = 0)

in_rc   <- recodes(lag1_A  = A,   # At each subsequent step, copy current values
                   lag1_L1 = L1,
                   lag1_L2 = L2)
```

If lag columns are already present in the data (pre-computed before calling `gformula()`), you do not need recoding hooks.

---

## Total Effect Estimation with `gformula()`

### Example 1 — Binary end-of-follow-up outcome

```{r gformula-binary}
# ── 1. Specify models in temporal order: L1 → L2 → A → Y ──────────────────
m_L1 <- spec_model(L1    ~ lag1_A + lag1_L1 + V + time,
                   var_type = "normal",  mod_type = "covariate")
m_L2 <- spec_model(L2    ~ lag1_A + lag1_L2 + V + time,
                   var_type = "binary",  mod_type = "covariate")
m_A  <- spec_model(A     ~ lag1_A + L1 + L2 + V + time,
                   var_type = "binary",  mod_type = "exposure")
m_Y  <- spec_model(Y_bin ~ A + L1 + L2,
                   var_type = "binary",  mod_type = "outcome")

models_bin <- list(m_L1, m_L2, m_A, m_Y)

# ── 2. Define intervention strategies ──────────────────────────────────────
# NULL  = natural course (draw exposure from its fitted model)
# 1 / 0 = always treat / never treat
ints <- list(natural = NULL, always_treat = 1, never_treat = 0)

# ── 3. Run g-formula ────────────────────────────────────────────────────────
fit_bin <- gformula(
  data        = nonsurvivaldata,
  id_var      = "id",
  time_var    = "time",
  base_vars   = "V",
  exposure    = "A",
  models      = models_bin,
  intervention = ints,
  ref_int     = "natural",
  init_recode = init_rc,
  in_recode   = in_rc,
  mc_sample   = 10000,
  R           = 1,        # set R > 1 for bootstrap CIs; kept low here for speed
  quiet       = TRUE,
  seed        = 2025
)
```

```{r gformula-binary-results}
# Risk (mean outcome) under each strategy
fit_bin$effect_size

# Contrasts vs the reference (natural course)
fit_bin$estimate
```

The `effect_size` table gives the estimated mean outcome under each intervention. The `estimate` table gives contrasts (risk difference and risk ratio) against the reference arm.

### Example 2 — Survival (time-to-event) outcome

For survival outcomes use `mod_type = "survival"`. The model estimates the discrete-time hazard at each time point; the package accumulates these into cumulative incidence (1 − product-limit survival).

```{r gformula-survival}
data("survivaldata")

m_L  <- spec_model(L ~ V + lag1_L + time,
                   var_type = "normal", mod_type = "covariate")
m_A2 <- spec_model(A ~ V + lag1_A + lag1_L + L + time,
                   var_type = "binary", mod_type = "exposure")
m_Y2 <- spec_model(Y ~ lag1_A + A + L + lag1_L + time,
                   var_type = "binary", mod_type = "survival")  # <-- survival

models_surv <- list(m_L, m_A2, m_Y2)

fit_surv <- gformula(
  data        = survivaldata,
  id_var      = "id",
  base_vars   = "V",
  exposure    = "A",
  time_var    = "time",
  models      = models_surv,
  intervention = list(natural = NULL, never = 0, always = 1),
  ref_int     = "natural",
  init_recode = recodes(lag1_L = 0, lag1_A = 0),
  in_recode   = recodes(lag1_L = L, lag1_A = A),
  mc_sample   = 10000,
  R           = 1,
  quiet       = TRUE,
  seed        = 2025
)

fit_surv$effect_size   # Cumulative incidence by strategy
fit_surv$estimate      # Risk contrasts
```

### Example 3 — Dynamic (threshold) intervention

A **dynamic intervention** assigns exposure based on a rule that depends on the individual's current covariate values. For example, we might treat individuals when a covariate exceeds a specified threshold. Wrap the rule in `dyn_int()`:

```{r gformula-dynamic}
fit_dyn <- gformula(
  data        = nonsurvivaldata,
  id_var      = "id",
  time_var    = "time",
  base_vars   = "V",
  exposure    = "A",
  models      = models_bin,
  intervention = list(
    natural  = NULL,
    treat_if_L1_pos = dyn_int(as.numeric(L1 > 0))
  ),
  ref_int     = "natural",
  init_recode = init_rc,
  in_recode   = in_rc,
  mc_sample   = 10000,
  R           = 1,
  quiet       = TRUE,
  seed        = 2025
)

fit_dyn$effect_size
fit_dyn$estimate
```

The expression inside `dyn_int()` is evaluated within the simulated dataset at each time step. All current column values — including the exposure after its natural-course draw — are in scope, so rules like `dyn_int(as.numeric(A > 0 & L1 > median(L1)))` are valid.

---

## Causal Mediation Analysis with `mediation()`

The `mediation()` function estimates the causal effect decomposition:

$$\underbrace{E[Y_{1,M(1)}] - E[Y_{0,M(0)}]}_{\text{Total effect}} =
  \underbrace{E[Y_{1,M(0)}] - E[Y_{0,M(0)}]}_{\text{Direct effect}} +
  \underbrace{E[Y_{1,M(1)}] - E[Y_{1,M(0)}]}_{\text{Indirect effect}}$$

where $Y_{a, M(a^*)}$ denotes the potential outcome under exposure $a$ with the mediator set to what it would have been under $a^*$.

You must include a model with `mod_type = "mediator"` in the model list, and pass the outcome variable name via `outcome`.

### Requirements and defaults

**Binary exposure.** The exposure variable must be coded as `0` (reference/untreated) and `1` (active/treated). Both Lin et al. (2017) and Zheng & van der Laan (2017) are defined for binary exposures; `mediation()` will stop with an informative error if other values are found.

**Single mediator.** Exactly one model with `mod_type = "mediator"` is required.

**Temporal ordering.** The list order determines the simulation sequence. Two orderings are common:

- `A → L → M → S` — confounders at time *t* are not affected by the mediator at time *t* (mediator conditions on L; outcome conditions on A, M, L).
- `A → M → L → S` — confounders at time *t* are affected by both exposure and mediator (Lin et al. 2017 DAG). This is the setting where standard regression fails.

The function checks that exposure precedes the mediator and that the mediator precedes the outcome, and warns if either is violated.

**Survivor-restricted mediator pool (interventional effects, survival outcomes).** For `mediation_type = "I"`, mediator values under `a* = 0` are drawn using weighted sampling based on the cumulative survival probability (`Sc`) at each time point, thereby implementing the survivor conditioning required by Lin et al. (2017, Eq. 4).

**Warning summary.** Warnings from model fitting (e.g., convergence, near-separation) are held and printed as a deduplicated summary at function exit. Repeated warnings (e.g., across 500 bootstrap replicates) are shown once with a count.

### Setting up models for mediation

The mediator model must condition on exposure and all confounders at the same time point; the outcome model must include both exposure and mediator. The example below uses the `A → L → M → Y` ordering (confounders not affected by the mediator).

```{r mediation-models}
init_med <- recodes(lag1_A = 0, lag1_L1 = 0, lag1_L2 = 0, lag1_M = 0)
in_med   <- recodes(lag1_A = A, lag1_L1 = L1, lag1_L2 = L2, lag1_M = M)

models_med <- list(
  spec_model(A   ~ V + lag1_L1 + lag1_L2 + lag1_A + time,
             var_type = "binary",  mod_type = "exposure"),
  spec_model(L1  ~ V + A + lag1_L1 + time,
             var_type = "normal",  mod_type = "covariate"),
  spec_model(L2  ~ V + A + lag1_L2 + time,
             var_type = "binary",  mod_type = "covariate"),
  spec_model(M   ~ V + A + L1 + L2 + lag1_M + time,
             var_type = "normal",  mod_type = "mediator"),   # <-- mediator
  spec_model(Y_bin ~ V + A + M + L1 + L2,
             var_type = "binary",  mod_type = "outcome")
)
```

### Interventional effects (IDE/IIE) — Lin et al. (2017)

Interventional effects use the **marginal** mediator distribution: the mediator is drawn by random permutation from values simulated under the reference exposure $a^* = 0$, breaking any individual-level dependence between mediator and confounders. This approach does not require cross-world independence and is generally preferred for time-varying settings.

```{r mediation-interventional}
fit_ide <- mediation(
  data           = nonsurvivaldata,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y_bin",
  models         = models_med,
  init_recode    = init_med,
  in_recode      = in_med,
  mediation_type = "I",     # Interventional IDE/IIE
  mc_sample      = 10000,
  R              = 1,
  quiet          = TRUE,
  seed           = 2025
)
```

```{r mediation-interventional-results}
# Intermediate counterfactual risks: Q(1,1), Q(1,0), Q(0,0)
fit_ide$effect_size

# Effect decomposition
fit_ide$estimate
```

### Natural effects (NDE/NIE) — Zheng & van der Laan (2017)

Natural effects use the **conditional** mediator distribution: at each time step the mediator is predicted using the individual's own covariate history under $a = 1$ but with the exposure argument in the mediator model swapped to $a^* = 0$. This requires the additional assumption of no unmeasured exposure-induced mediator–outcome confounding.

```{r mediation-natural}
fit_nde <- mediation(
  data           = nonsurvivaldata,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y_bin",
  models         = models_med,
  init_recode    = init_med,
  in_recode      = in_med,
  mediation_type = "N",     # Natural NDE/NIE
  mc_sample      = 10000,
  R              = 1,
  quiet          = TRUE,
  seed           = 2025
)

fit_nde$estimate
```

The key identifiability difference between `"I"` and `"N"`:

| | Interventional (`"I"`) | Natural (`"N"`) |
|--|----------------------|-----------------|
| Mediator drawn from | Marginal distribution (permutation) | Conditional distribution (individual history) |
| Cross-world independence required | No | Yes |
| Interpretation | Shift in mediator *population* distribution | Hypothetical individual-level swap |
| Reference | Lin et al. (2017) | Zheng & van der Laan (2017) |

---

## Bootstrap Confidence Intervals

Set `R > 1` to obtain percentile and normal-approximation confidence intervals. The bootstrap resamples whole subjects (all time points together), preserving the longitudinal correlation structure.

```{r bootstrap, eval=FALSE}
fit_boot <- gformula(
  data        = nonsurvivaldata,
  id_var      = "id",
  time_var    = "time",
  base_vars   = "V",
  exposure    = "A",
  models      = models_bin,
  intervention = list(natural = NULL, always = 1),
  ref_int     = "natural",
  init_recode = init_rc,
  in_recode   = in_rc,
  mc_sample   = 10000,
  R           = 200,       # 200 bootstrap replicates
  quiet       = TRUE,
  seed        = 2025
)

# effect_size now includes Sd, perct_lcl/ucl, norm_lcl/ucl
fit_boot$effect_size
fit_boot$estimate
```

### Parallel bootstrap

The bootstrap loop uses `future.apply::future_lapply` internally. Enable parallelism by setting a parallel plan before calling the function:

```{r parallel, eval=FALSE}
library(future)
plan(multisession)   # use multiple R sessions in parallel

fit_par <- gformula(..., R = 500)

plan(sequential)     # restore default after use
```

On Unix/macOS, `plan(multicore)` (forking) is more memory-efficient.

---

## Working with Results

### Extracting fitted models

Set `return_fitted = TRUE` to access the full fitted model objects and their coefficients:

```{r fitted-models}
fit_full <- gformula(
  data        = nonsurvivaldata,
  id_var      = "id",
  time_var    = "time",
  base_vars   = "V",
  exposure    = "A",
  models      = models_bin,
  intervention = list(natural = NULL, always = 1),
  init_recode = init_rc,
  in_recode   = in_rc,
  mc_sample   = 5000,
  R           = 1,
  return_fitted = TRUE,
  quiet       = TRUE,
  seed        = 2025
)

# Names correspond to the response variable of each model
names(fit_full$fitted_models)

# Access a specific model's coefficients
coef(fit_full$fitted_models$A)
```

### Retrieving the simulated data

Set `return_data = TRUE` to retrieve the full Monte Carlo dataset (can be large):

```{r return-data, eval=FALSE}
fit_data <- gformula(..., return_data = TRUE)

# Simulated trajectories with predicted outcomes, one row per MC subject per time
head(fit_data$sim_data)
```

---

## Causal Assumptions

Valid inference requires the following assumptions to hold for the observed data:

1. **Consistency**: the potential outcome under the observed exposure history equals the observed outcome.
2. **Positivity**: every covariate pattern that occurs under the intervention also occurs in the observed data (non-zero probability of receiving each exposure level).
3. **Sequential exchangeability**: no unmeasured confounding of the exposure–outcome relationship at each time point, conditional on the measured past.

For **natural effects** (`mediation_type = "N"`), an additional assumption is required:

4. **No unmeasured exposure-induced mediator–outcome confounding**: there are no confounders of the mediator–outcome relationship that are themselves caused by prior exposure. Interventional effects (`mediation_type = "I"`) do not require this assumption.

---

## Non-standard Covariate Distributions

The four built-in `var_type` values cover the most common cases:

| `var_type` | Distribution |
|---|---|
| `"binary"` | Bernoulli (logistic regression) |
| `"normal"` | Gaussian (linear regression) |
| `"categorical"` | Multinomial (`nnet::multinom`) |
| `"custom"` | User-supplied fit and simulation functions |

For distributions not in this list — **bounded normal**, **truncated normal**, **zero-inflated normal**, **absorbing states** — use `var_type = "custom"` with the `custom_fit` and `custom_sim` arguments to `spec_model()`:

```{r custom-type, eval=FALSE}
# Example: bounded normal covariate in [0, 1]
my_fit <- function(formula, data, ...) {
  # fit any model you like; return the fitted object
  lm(formula, data = data)
}

my_sim <- function(model, newdt, ...) {
  pred <- predict(model, newdata = newdt)
  # draw from N(pred, sigma^2) and clamp to [0, 1]
  draws <- rnorm(length(pred), pred, sigma(model))
  pmin(pmax(draws, 0), 1)
}

m_bounded <- spec_model(
  X ~ A + L + time,
  var_type   = "custom",
  mod_type   = "covariate",
  custom_fit = my_fit,
  custom_sim = my_sim
)
```

`custom_fit(formula, data, ...)` is called during model fitting and must return an object that `custom_sim` knows how to predict from. `custom_sim(model, newdt, ...)` is called at each simulation step and must return a vector of simulated values of length `nrow(newdt)`.

---

## Reference

- Westreich, D., Cole, S. R., Young, J. G., et al. (2012). The parametric g-formula to estimate the effect of highly active antiretroviral therapy on incident AIDS or death. *Statistics in Medicine*, 31, 2000–2009.
- McGrath, S., Lin, V., Zhang, Z., et al. (2020). gfoRmula: An R package for estimating the effects of sustained treatment strategies via the parametric g-formula. *Patterns*, 1, 100008.
- Zheng, W., & van der Laan, M. (2017). Longitudinal mediation analysis with time-varying mediators and exposures, with application to survival outcomes. *Journal of Causal Inference*, 5(2), 20160006.
- Lin, S.-H., Young, J. G., Logan, R., & VanderWeele, T. J. (2017). Mediation analysis for a survival outcome with time-varying exposures, mediators, and confounders. *Statistics in Medicine*, 36, 4153–4166.
