---
title: "Mediation Analysis for Survival Outcomes"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    highlight: null
vignette: >
  %\VignetteIndexEntry{Mediation Analysis for Survival Outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Introduction

The mediational g-formula of Lin et al. (2017) was developed for **survival
outcomes** with time-varying exposures, mediators, and confounders. This
vignette walks through that setting:

1. a single-mediator interventional analysis on a survival outcome,
2. how to read every row of the output,
3. the `n_vw` permutation-averaging parameter,
4. multiple mediators — replicating the Yamamuro et al. (2021) simulation
   setting, with estimates checked against documented true values,
5. censoring, and
6. restricting models with `subset` (absorbing states).

For an introduction to the package (model specification, recode hooks,
`gformula()` for total effects) see `vignette("causalMed-overview")`.

```{r load}
library(causalMed)
library(data.table)
```

---

## Survival Data Requirements

With a survival outcome, `mediation()` (and `gformula()`) model the
**discrete-time hazard**: at each time point, a binary model with
`mod_type = "survival"` estimates the probability of the event among those
still at risk. The simulation accumulates these hazards into a cumulative
incidence, $1 - \prod_t (1 - h_t)$, so all reported quantities — the per-arm
estimates and the effect decomposition — are **risks of the event by the end
of follow-up**.

The data must be in long format with one row per subject per period **at
risk**: once the event occurs, no later rows for that subject may be present.

```{r data}
data("survivaldata")
dat <- as.data.table(survivaldata)
head(dat, 8)
```

`survivaldata` contains 3 000 subjects followed over five periods
(`time` = 0–4):

| Variable | Role |
|----------|------|
| `id` | Subject identifier |
| `time` | Time index |
| `V` | Time-fixed baseline covariate |
| `A` | Time-varying binary exposure |
| `L` | Time-varying continuous confounder |
| `M` | Time-varying binary mediator |
| `Y` | Event indicator (1 = event in this period) |
| `lag1_A`, `lag1_L`, `lag1_M` | Previous-period values |

The data-generating ordering within each period is **A → L → M → Y**
(see `?survivaldata`): exposure first, then the confounder (affected by
current exposure), then the mediator (affected by current exposure and
confounder), then the hazard.

---

## Single Mediator: Interventional Effects

Models are listed in the temporal order above. Note `mod_type = "survival"`
for the outcome; the formula may include exposure–mediator interaction terms.

```{r models}
init_s <- recodes(lag1_A = 0, lag1_L = 0, lag1_M = 0)
in_s   <- recodes(lag1_A = A, lag1_L = L, lag1_M = M)

models_surv <- list(
  spec_model(A ~ V + lag1_A + lag1_L + time,
             var_type = "binary", mod_type = "exposure"),
  spec_model(L ~ V + A + lag1_L + time,
             var_type = "normal", mod_type = "covariate"),
  spec_model(M ~ V + A + L + lag1_M + time,
             var_type = "binary", mod_type = "mediator"),
  spec_model(Y ~ V + A + M + L + A:M + time,
             var_type = "binary", mod_type = "survival")   # discrete-time hazard
)
```

```{r fit-single}
fit_surv <- mediation(
  data           = dat,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y",
  models         = models_surv,
  init_recode    = init_s,
  in_recode      = in_s,
  mediation_type = "I",
  mc_sample      = 20000,
  R              = 1,      # set R > 1 for bootstrap CIs (see below)
  quiet          = TRUE,
  seed           = 2026
)
```

### Reading the per-arm table

```{r arms-single}
fit_surv$effect_size
```

Each row is the simulated cumulative incidence under one arm:

- **`nat0` / `nat1`** — exposure fixed to 0 / 1, mediator following its
  fitted model given each individual's own history. Their contrast is the
  natural plug-in **total effect**.
- **`Phi00` / `Phi11`** — exposure fixed to 0 / 1, mediator drawn from the
  *permuted marginal pool* collected under that same exposure level
  ($E[Y_{0,G_0}]$, $E[Y_{1,G_1}]$). These are the interventional reference
  arms.
- **`Phi10`** — the cross-world arm: exposure fixed to 1, mediator drawn from
  the a = 0 pool ($E[Y_{1,G_0}]$).

### Reading the decomposition

```{r decomp-single}
fit_surv$estimate
```

Row by row:

- **Indirect effect (IIE)** = `Phi11 − Phi10`: the change in risk from
  shifting the *population distribution* of the mediator from its
  never-treated to its always-treated form, while exposure is held at 1.
- **Direct effect (IDE)** = `Phi10 − Phi00`: the effect of exposure with the
  mediator distribution held at its never-treated form.
- **Total effect (TE)** = `nat1 − nat0`: the ordinary g-formula total effect.
- **TE − (Direct + Indirect)**: IDE + IIE sum to the *interventional overall
  effect* `Phi11 − Phi00`, not to TE; this row is the difference (a
  mediated-interaction residual). It is usually small; a large value signals
  strong exposure–mediator interaction in the outcome process. (For
  `mediation_type = "N"` the decomposition is exact and this row is absent.)
- **Mediation Proportion** = (TE − IDE) / TE × 100: the share of the total
  effect *not* acting directly, i.e. the indirect effects plus the residual
  (Yamamuro et al. 2021).
- **Mediation Proportion (multiplicative)** =
  $RR_{IDE}(RR_{IIE}-1)/(RR_{OE}-1) \times 100$ on the risk-ratio scale
  (Lin et al. 2017, Table 2).

The `RD` column is the risk-difference scale, `RR` the risk-ratio scale
(`RR` is not applicable to the residual and proportion rows). With `R > 1`
both scales get bootstrap standard errors and percentile/normal confidence
intervals.

`print(fit_surv)` displays both tables together with the arm definitions as
a legend.

### The `n_vw` argument

Every cross-world arm draws the mediator by randomly permuting the pool of
simulated mediator trajectories. `n_vw` controls how many independent
permutations are averaged per arm (default `2`, matching the SAS
`mGFORMULA` macro). Averaging reduces Monte Carlo noise from the permutation
step at the cost of one extra simulation pass per arm; `n_vw = 1` is faster
and slightly noisier. It has no effect on natural effects
(`mediation_type = "N"`), which do not use permutation.

```{r nvw, eval=FALSE}
mediation(..., n_vw = 1)   # single permutation per arm: faster, noisier
```

---

## Multiple Mediators: the Yamamuro et al. (2021) Simulation

With `mediation_type = "I"`, any number of mediators can be analysed by
supplying several `mod_type = "mediator"` models **in temporal order**. We
demonstrate this on `yamamurodata`, a dataset simulated from the
data-generating process of the Yamamuro et al. (2021) simulation study —
a time-varying treatment `A`, a confounder `L`, two sequential mediators
`M1` and `M2`, and a survival outcome over three visits, ordered
**A → L → M1 → M2 → Y** within each visit. Because the process is known,
the **true interventional effects are documented in `?yamamurodata`**
(computed at $n = 10^7$), so the estimates below can be checked against
the truth.

```{r yam-data}
data("yamamurodata")
yam <- as.data.table(yamamurodata)
head(yam, 6)
```

Three specification details are worth noting:

- **Visit indicators.** The published models give each visit its own
  intercept. Write these as `I(as.integer(time == k))` rather than
  `factor(time)`: during simulation each Monte Carlo time slice carries a
  single `time` value, so a factor would drop the unobserved levels while
  the numeric indicator predicts safely at every step.
- **`subset = time > 0`.** The visit-0 values of `A`, `L`, `M1`, `M2` are
  baseline draws, not model output, so the time-varying models are fitted
  and simulated only for `time > 0`. Visit 0 is seeded in the Monte Carlo
  cohort from the time-fixed baseline columns (`L0base`, `M10base`,
  `M20base`) via `init_recode`.
- The models include the quadratic and lag terms of the published
  "correctly specified" scenario.

```{r yam-models}
models_yam <- list(
  spec_model(A ~ V + lag1_A + lag1_L + I(lag1_L^2) + lag1_M1 + lag1_M2 +
               I(as.integer(time == 2)),
             var_type = "binary", mod_type = "exposure", subset = time > 0),
  spec_model(L ~ V + A + lag1_L + I(lag1_L^2) + lag1_M1 + lag1_M2 +
               I(as.integer(time == 2)),
             var_type = "normal", mod_type = "covariate", subset = time > 0),
  spec_model(M1 ~ V + A + L + I(L^2) + lag1_L + I(lag1_L^2) + lag1_M1 + lag1_M2 +
               I(as.integer(time == 2)),
             var_type = "normal", mod_type = "mediator", subset = time > 0),
  spec_model(M2 ~ V + A + L + I(L^2) + lag1_L + I(lag1_L^2) + lag1_M1 + lag1_M2 +
               I(as.integer(time == 2)),
             var_type = "normal", mod_type = "mediator", subset = time > 0),
  spec_model(Y ~ V + A + L + I(L^2) + M1 + M2 + M1:M2 +
               I(as.integer(time == 1)) + I(as.integer(time == 2)),
             var_type = "binary", mod_type = "survival")
)

fit_yam <- mediation(
  data           = yam,
  id_var         = "id",
  time_var       = "time",
  base_vars      = c("V", "L0base", "M10base", "M20base"),
  exposure       = "A",
  outcome        = "Y",
  models         = models_yam,
  init_recode    = recodes(L = L0base, M1 = M10base, M2 = M20base,
                           lag1_A = 0, lag1_L = 0, lag1_M1 = 0, lag1_M2 = 0),
  in_recode      = recodes(lag1_A = A, lag1_L = L, lag1_M1 = M1, lag1_M2 = M2),
  mediation_type = "I",
  n_vw           = 2,      # matches the SAS mGFORMULA macro
  mc_sample      = 20000,
  R              = 1,
  quiet          = TRUE,
  seed           = 2026
)

fit_yam$effect_size
```

For $N$ mediators the arm list grows to $4 + N$: the intermediate arms
`Phi1_k` switch the first $k$ mediators to the a = 1 pool while the rest
stay at the a = 0 pool. Each mediator's indirect effect is the **sequential
contrast** $IIE(M_k) = \Phi_{1,k} - \Phi_{1,k-1}$, labelled
`Indirect effect (<name>)` in the output, and
$IDE + \sum_k IIE(M_k)$ equals the interventional overall effect
`Phi11 − Phi00`. Each mediator's pool is permuted independently.

### Comparing against the true values

```{r yam-truth}
truth <- data.table(
  Effect = c("Total effect", "Direct effect", "Indirect effect (M1)",
             "Indirect effect (M2)", "TE - (Direct + Indirect)"),
  True   = c(-6.36, -3.20, -2.29, -0.97, 0.10)   # from ?yamamurodata
)
est <- as.data.table(fit_yam$estimate)[, .(Effect, Estimate = RD * 100)]
merge(truth, est, by = "Effect", sort = FALSE)
```

The estimates reproduce the structure of the truth: all signs and the
relative magnitudes (IDE > IIE via M1 > IIE via M2, small positive
residual) are recovered, and the two indirect effects are estimated
closely. The total and direct effects deviate by roughly 1–1.5 percentage
points — sampling error from fitting on a single dataset. The published
study averages 1000 replicate datasets, under which the mean estimates
match the true values (Yamamuro et al. 2021, and reproduced with this
package during development); a per-dataset bootstrap (`R > 1`) quantifies
this uncertainty in applied use.

Because the decomposition is sequential, the *order* of the mediator models
matters: it must reflect the assumed temporal/causal ordering among the
mediators. Multiple mediators are not available for `mediation_type = "N"`
(the natural-effects references define a single mediator only); `mediation()`
stops with an error in that case.

---

## Censoring

If follow-up can end for reasons other than the event, include a censoring
indicator with `mod_type = "censor"`. To illustrate, we censor some
follow-up in `survivaldata` with probability depending on the confounder
(informative censoring), ending each subject's follow-up at their first
censoring:

```{r censor-data}
set.seed(11)
dat_c <- copy(dat)
dat_c[, C := rbinom(.N, 1, plogis(-3.5 + 0.8 * L))]
dat_c[, after_cens := cumsum(shift(C, fill = 0)), by = id]
dat_c <- dat_c[after_cens == 0][, after_cens := NULL]
dat_c[C == 1, Y := 0L]   # censored before the event in that period
```

Add a censoring model to the list (after the covariates/mediator it depends
on, before the survival model):

```{r fit-censor}
models_cens <- list(
  spec_model(A ~ V + lag1_A + lag1_L + time,
             var_type = "binary", mod_type = "exposure"),
  spec_model(L ~ V + A + lag1_L + time,
             var_type = "normal", mod_type = "covariate"),
  spec_model(M ~ V + A + L + lag1_M + time,
             var_type = "binary", mod_type = "mediator"),
  spec_model(C ~ V + A + L + time,
             var_type = "binary", mod_type = "censor"),     # censoring model
  spec_model(Y ~ V + A + M + L + A:M + time,
             var_type = "binary", mod_type = "survival")
)

fit_cens <- mediation(
  data           = dat_c,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y",
  models         = models_cens,
  init_recode    = init_s,
  in_recode      = in_s,
  mediation_type = "I",
  mc_sample      = 20000,
  R              = 1,
  quiet          = TRUE,
  seed           = 2026
)

fit_cens$estimate
```

In the simulation, censoring is **abolished** in every arm (all arms fix the
exposure, and under an intervention the censoring indicator is set to 0), so
the reported risks are counterfactual risks *in the absence of censoring* —
the standard g-formula treatment of right-censoring. The censoring model's
role is to let the hazard model be fitted on data where censoring depends on
measured covariates. Here the estimates from the censored data are close to
the full-data estimates above, as expected when censoring depends only on
modelled covariates.

---

## Restricting Models with `subset` (Absorbing States)

`spec_model(subset = ...)` fits and simulates a model only on rows meeting a
condition. The classic use is an **absorbing state** — for example, in the
GvHD analysis of Westreich et al. (2012) the exposure can only switch on
once, so its model is estimated among the not-yet-exposed and the value is
carried forward deterministically afterwards:

```{r subset, eval=FALSE}
# exposure can occur only while gvhdm1 == 0 …
spec_model(gvhd ~ all + cmv + male + age + ...,
           var_type = "binary", mod_type = "exposure",
           subset = gvhdm1 == 0)

# … and is locked at 1 afterwards via the end-of-step hook
out_recode = recodes(gvhd = ifelse(gvhdm1 == 1, 1, gvhd))
```

Rows excluded by `subset` keep their current value at that step, so the
`out_recode` carry-forward completes the absorbing behaviour.

---

## Bootstrap Confidence Intervals

As elsewhere in the package, set `R > 1` for subject-level bootstrap CIs and
optionally register a parallel plan first:

```{r boot, eval=FALSE}
library(future)
plan(multisession)

fit_ci <- mediation(
  data           = dat,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y",
  models         = models_surv,
  init_recode    = init_s,
  in_recode      = in_s,
  mediation_type = "I",
  mc_sample      = 20000,
  R              = 500,
  seed           = 2026
)

plan(sequential)

# estimate now carries Sd, percentile and normal CIs on the RD and RR scales
fit_ci$estimate
```

---

## Natural Effects and Intermediate Confounding

`mediation_type = "N"` (Zheng & van der Laan 2017) is also defined for
survival outcomes. However, in this dataset the confounder `L` is affected by
the current exposure — an *intermediate confounder* — and natural direct and
indirect effects are **not identifiable** in that setting. `mediation()`
detects the exposure on the right-hand side of a covariate model and warns:

```{r natural, warning=TRUE}
fit_nat <- mediation(
  data           = dat,
  id_var         = "id",
  time_var       = "time",
  base_vars      = "V",
  exposure       = "A",
  outcome        = "Y",
  models         = models_surv,
  init_recode    = init_s,
  in_recode      = in_s,
  mediation_type = "N",
  mc_sample      = 20000,
  R              = 1,
  quiet          = TRUE,
  seed           = 2026
)
```

The interventional effects (`"I"`) remain identifiable under intermediate
confounding and are the recommended estimand here — which is why they are
the package default.

---

## References

- Lin, S.-H., Young, J. G., Logan, R., & VanderWeele, T. J. (2017). Mediation
  analysis for a survival outcome with time-varying exposures, mediators, and
  confounders. *Statistics in Medicine*, 36, 4153–4166.
- VanderWeele, T. J., & Tchetgen Tchetgen, E. J. (2017). Mediation analysis
  with time varying exposures and mediators. *Journal of the Royal
  Statistical Society: Series B*, 79(3), 917–938.
- Yamamuro, S., Shinozaki, T., Iimuro, S., & Matsuyama, Y. (2021).
  Mediational g-formula for time-varying treatment and repeated-measured
  multiple mediators. *Statistical Methods in Medical Research*, 30(8),
  1782–1799.
- Zheng, W., & van der Laan, M. (2017). Longitudinal mediation analysis with
  time-varying mediators and exposures, with application to survival
  outcomes. *Journal of Causal Inference*, 5(2).
- Westreich, D., Cole, S. R., Young, J. G., et al. (2012). The parametric
  g-formula to estimate the effect of highly active antiretroviral therapy on
  incident AIDS or death. *Statistics in Medicine*, 31, 2000–2009.
